---
title: "Installation details"
lang: en-GB
output:
  rmarkdown::html_vignette:
    mathjax: null
    includes:
      in_header: vignette-layout.inc
    toc: true
vignette: >
  %\VignetteIndexEntry{Installation details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

For the ordinary installation, start with [Install lisaR](installation.html).
This page keeps the optional dependencies and installation diagnostics.
You need **R 4.1 or later**. RStudio is optional, but the steps below work in
its Console.

If you only need to read a portable HTML report, you do not need R. Keep the
report folder and its relative files together.

lisaR is installed from its GitHub repository. The installation includes the
LISA dictionaries and the synthetic sample project used in Quick Start.

## Install the required packages

Run this once in R with internet access. `BiocManager` installs `fgsea` from a
Bioconductor release that is compatible with your R version, together with the
required CRAN packages.

```{r installation-dependencies, eval=FALSE}
install.packages(c("BiocManager", "remotes"))
BiocManager::install(
  c("digest", "fgsea", "fs", "ggplot2", "hommel", "jsonlite", "openxlsx", "readr", "yaml"),
  update = FALSE,
  ask = FALSE
)
```

The required `hommel` version is at least 1.8; the installer resolves this from
`DESCRIPTION`. Base and recommended Imports come with R. Install extra packages
only for the route you intend to use:

| Route | Additional packages |
|---|---|
| Shiny application with background figure jobs | `shiny`, `callr`, `ps` |
| Prepared Riaz project | `data.table` |
| Gene-level figures and FULL cards | `ggrepel`, `gridExtra`, `patchwork` |
| Native KEGG maps | `AnnotationDbi`, `KEGGREST`, `png`, plus `org.Hs.eg.db` for human or `org.Mm.eg.db` for mouse |
| Reconstructing Riaz upstream inputs | `data.table`, `readxl`, `DESeq2`, `AnnotationDbi`, `org.Hs.eg.db`, `SummarizedExperiment` |
| Building these HTML vignettes from source | `knitr`, `rmarkdown`, plus Pandoc outside R |

Use the same `BiocManager::install()` form for these packages. For example,
install `data.table` before using the prepared Riaz project. Later,
`validate_lisa_config()` will report packages needed by the analysis and
report products you selected.

## Install lisaR from GitHub

In R or the RStudio Console, run:

```{r install-github, eval=FALSE}
remotes::install_github(
  "DBM-OlmedaLab/lisaR",
  upgrade = "never",
  build_vignettes = FALSE
)
```

This installs the package and its included resources in one step. You do not
need to locate a separate source archive or install lisaR again to load its
dictionaries. Ordinary installation does not rebuild the vignettes.

The source package includes the LISA dictionaries, category hierarchy and
synthetic sample project. It does **not** include the external MSigDB
TERM2GENE membership needed for a real study. Preparing that membership is a
separate step in [Resources and provenance](resources-and-provenance.html).

Source and release status are listed at
[DBM-OlmedaLab/lisaR](https://github.com/DBM-OlmedaLab/lisaR). Repository access
may remain restricted before publication. If GitHub reports that the repository
is unavailable, check your access with the project owner. Do not paste access
tokens into a chat or into this code.

## Check that R loads the installed copy

Restart R if another lisaR version is already loaded, then run:

```{r confirm-installation, eval=FALSE}
library(lisaR)
packageVersion("lisaR")
find.package("lisaR")
system.file("extdata", "dictionaries", package = "lisaR")
system.file("examples", "quick-start", "study.yml", package = "lisaR")
```

`packageVersion()` identifies the installed package. `find.package()` shows
the R library from which it was loaded. If these checks complete, continue
with [Quick Start](quick-start.html). That synthetic project contains every
input needed for a first report and does not require MSigDB access.

For a real study, continue to
[Prepare differential-expression inputs](preparing-de-inputs.html), then
[run lisaR with your prepared files](getting-started.html).

## Read the guides

The installation above does not rebuild HTML vignettes. Read the
[guides online](https://olmedalab.org/lisaR/en/#guides), where each guide also
has a PDF download. You can open the Quick Start directly from R:

```{r open-online-guide, eval=FALSE}
browseURL("https://olmedalab.org/lisaR/reader/quick-start.html?lang=en")
```

The source files are in the repository's `vignettes/` directory, in `.Rmd`
format. `vignette()` lists compiled HTML only when the package was built with
vignettes. You do not need those HTML files or another installation to run the
examples.

## If installation fails

| Message or problem | Action |
|---|---|
| A package dependency is missing | Run the dependency step in the same R library, then install lisaR again. |
| R and Bioconductor versions are incompatible | Select the Bioconductor release supported by the active R version. |
| A source dependency requires compilation | Install the platform's R build tools: Rtools on Windows, the command-line tools on macOS, or the appropriate compiler packages on Linux. |
| Pandoc is missing during a vignette build | Use the Pandoc supplied by RStudio or install it before rebuilding documentation. Pandoc is not needed for an ordinary lisaR analysis. |
| An older lisaR version still appears | Restart R and check the library returned by `find.package("lisaR")`. |

See [Troubleshooting](troubleshooting.html) for configuration and execution
errors after installation.
