---
title: "Custom dictionaries and supercategories"
lang: en-GB
output:
  rmarkdown::html_vignette:
    mathjax: null
    includes:
      in_header: vignette-layout.inc
    toc: true
vignette: >
  %\VignetteIndexEntry{Custom dictionaries and supercategories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A custom dictionary lets you organise gene sets using a classification that
fits your study. It is useful when a specialised assay or a laboratory-curated
collection is not represented well by the built-in LISA categories.

You need three related TSV files. TERM2GENE defines which genes belong to each
gene set. The dictionary assigns those gene sets to categories. The category
map names and orders the categories within broader supercategories. Register
the three files under new resource identities so that they do not replace the
built-in resources.

This guide follows one synthetic example from table construction to a saved
report. It reuses the 40-gene sample project included with lisaR, including
its gene-set memberships and differential-expression data. The custom category
assignments below are example labels, not a biological classification to
reuse.

## 1. Create a fresh sample project

Run the sections in order in an ordinary R or RStudio session. Start with a new
copy of the sample project. `lisa_init_project()` refuses to overwrite an
existing directory, so choose a new name if you have used this one before.

```{r create-project, eval=FALSE}
library(lisaR)

project <- lisa_init_project("custom")
resource_dir <- file.path(project, "custom-resources")
stopifnot(!dir.exists(resource_dir))
dir.create(resource_dir)
```

The copied project already contains the two synthetic analyses and their
40-gene input tables. The following sections construct only the custom
dictionary resources.

## 2. Construct the three resource tables

The sample TERM2GENE file contains 48 synthetic gene sets, each with 12
members drawn from `GENE01` to `GENE40`. Read those memberships and the
associated sample gene-set names from the installed package:

```{r construct-memberships, eval=FALSE}
sample_root <- system.file("extdata", "quick-start", package = "lisaR")
read_sample <- function(name) utils::read.delim(
  file.path(sample_root, name), sep = "\t", quote = "", comment.char = "",
  check.names = FALSE, stringsAsFactors = FALSE,
  colClasses = "character", na.strings = character()
)

term2gene <- read_sample("example_term2gene_v2_0.tsv")
sample_dictionary <- read_sample("lisa_dictionary_quickstart_v1_1.tsv")
gene_sets <- unique(sample_dictionary[, c(
  "universe", "gene_set_id", "gene_set_name"
)])

print(utils::head(term2gene, 12))
print(gene_sets)
```

Now define a new example classification. Each token selects one known family
of synthetic gene-set IDs. The category IDs, display names and two
supercategories are specific to this custom example.

```{r construct-classification, eval=FALSE}
classification <- data.frame(
  token = c("_SIGNAL_", "_STRESS_", "_PROLIF_", "_MATRIX_"),
  category_id = c(
    "LAB_SIGNAL", "LAB_STRESS", "LAB_PROLIFERATION", "LAB_MATRIX"
  ),
  category_display_name = c(
    "Example of signal response", "Example of stress response",
    "Example of proliferation", "Example of matrix remodelling"
  ),
  stringsAsFactors = FALSE
)

category_index <- vapply(gene_sets$gene_set_id, function(gene_set_id) {
  hits <- vapply(
    classification$token, grepl, logical(1),
    x = gene_set_id, fixed = TRUE
  )
  stopifnot(sum(hits) == 1L)
  which(hits)
}, integer(1))

dictionary <- data.frame(
  universe = gene_sets$universe,
  gene_set_id = gene_sets$gene_set_id,
  gene_set_name = gene_sets$gene_set_name,
  source_id = rep("CUSTOM_SAMPLE40", nrow(gene_sets)),
  category_id = classification$category_id[category_index],
  category_display_name =
    classification$category_display_name[category_index],
  tier = rep("custom", nrow(gene_sets)),
  stringsAsFactors = FALSE
)

category_map <- data.frame(
  category_id = classification$category_id,
  display_name = classification$category_display_name,
  macrogroup_id = c(
    "LAB_RESPONSES", "LAB_RESPONSES",
    "LAB_GROWTH_STRUCTURE", "LAB_GROWTH_STRUCTURE"
  ),
  macrogroup_name = c(
    "Example response programmes", "Example response programmes",
    "Example growth and structure", "Example growth and structure"
  ),
  macrogroup_order = c(1L, 1L, 2L, 2L),
  category_order_within_macrogroup = c(1L, 2L, 1L, 2L),
  stringsAsFactors = FALSE
)

stopifnot(
  setequal(dictionary$gene_set_id, term2gene$gs_name),
  all(dictionary$tier == "custom"),
  all(dictionary$category_id %in% category_map$category_id),
  all(dictionary$category_display_name ==
    category_map$display_name[match(
      dictionary$category_id, category_map$category_id
    )])
)

print(dictionary)
print(category_map)
```

These are the exact required fields:

| Table and schema | Required fields |
|---|---|
| TERM2GENE, `term2gene@1` | `gs_collection`, `gs_subcollection`, `gs_name`, `gs_exact_source`, `gene_symbol` |
| Dictionary, `lisa_dictionary@2` | `universe`, `gene_set_id`, `gene_set_name`, `source_id`, `category_id`, `category_display_name`, `tier` |
| Category map, `category_map@1` | `category_id`, `display_name`, `macrogroup_id`, `macrogroup_name`, `macrogroup_order`, `category_order_within_macrogroup` |

Use one TERM2GENE row for each exact gene-set and gene pair. Within a gene set,
keep `gs_collection`, `gs_subcollection` and `gs_exact_source` consistent.
`gene_symbol` must use the namespace expected for the declared species.

In the dictionary, `gene_set_id` must exactly match TERM2GENE `gs_name`. A gene
set may have more than one category assignment, but the same assignment must
not be repeated. Use `tier = "custom"`. The current schema does not have a
`LISA_score` column.

Each dictionary `category_id` needs one row in the category map, and
`category_display_name` must equal the map's `display_name`. A supercategory is
identified by `macrogroup_id` and named by `macrogroup_name`. Give each
supercategory a unique positive `macrogroup_order`, then number its categories
with unique positive `category_order_within_macrogroup` values.

## 3. Choose the collection for each gene set

The dictionary `universe` field accepts these semantic collections:

| Identifier | Scope |
|---|---|
| `GOBP-C2` | Broad biological programmes, combining GO Biological Process with selected curated pathway families from MSigDB C2:CP. |
| `GOMF` | GO Molecular Function gene sets. |
| `GOCC` | GO Cellular Component gene sets. |
| `PATHWAYS` | A strict signalling view of the GOBP-C2 gene-set space. |

`HALLMARKS` is not a custom dictionary universe. It is a direct TERM2GENE
collection and is analysed without semantic LISA remapping. A study can add
`HALLMARKS` only when its registered TERM2GENE resource contains compatible
`H` or `MH` rows.

## 4. Save and review the files

Save the three objects in the new project directory. These commands will not
silently replace an existing custom-resource directory because section 1
stops if that directory is already present.

```{r save-tables, eval=FALSE}
resource_files <- c(
  lisa_dictionary = file.path(resource_dir, "sample_dict.tsv"),
  category_map = file.path(resource_dir, "sample_map.tsv"),
  term2gene = file.path(resource_dir, "sample_t2g.tsv")
)

utils::write.table(
  dictionary, resource_files[["lisa_dictionary"]],
  sep = "\t", quote = FALSE, row.names = FALSE, na = ""
)
utils::write.table(
  category_map, resource_files[["category_map"]],
  sep = "\t", quote = FALSE, row.names = FALSE, na = ""
)
utils::write.table(
  term2gene, resource_files[["term2gene"]],
  sep = "\t", quote = FALSE, row.names = FALSE, na = ""
)
```

Stop before registration. Review the saved tables as files, not only the R
objects that created them. Confirm that all 48 gene sets have memberships,
that every dictionary category occurs in the map, and that the sample
species and pipeline profile are appropriate for this example. The sample
labels are synthetic and should not be presented as a reviewed classification
for another study.

## 5. Record the reviewed checksums in a manifest

Run this section only after the file review. It records the SHA-256 values of
the exact saved files in a persistent manifest. The review note is supplied by
the person performing the review rather than invented by the tutorial.

```{r create-manifest, eval=FALSE}
reviewer <- trimws(readline("Reviewer name or initials: "))
review_note <- trimws(readline("What did you check? "))
stopifnot(nzchar(reviewer), nzchar(review_note))

reviewed_at <- format(Sys.time(), "%Y-%m-%dT%H:%M:%S%z")
approved_origin <- paste0(
  "Local sample custom-resource review by ", reviewer,
  " on ", reviewed_at, ": ", review_note
)

reviewed_sha256 <- vapply(resource_files, function(path) {
  digest::digest(file = path, algo = "sha256", serialize = FALSE)
}, character(1))

manifest <- data.frame(
  logical_id = c(
    "my_lab_sample_dictionary",
    "my_lab_sample_category_map",
    "my_lab_sample_term2gene"
  ),
  version = rep("1.0.0", 3),
  species = rep("Homo sapiens", 3),
  modality = rep("transcriptomic/genomic", 3),
  schema = c("lisa_dictionary@2", "category_map@1", "term2gene@1"),
  sha256 = unname(reviewed_sha256),
  compatibility = rep("lisaR>=0.6.0", 3),
  approved_origin = rep(approved_origin, 3),
  artifact = basename(unname(resource_files)),
  stringsAsFactors = FALSE
)

manifest_path <- file.path(resource_dir, "resource_manifest.tsv")
utils::write.table(
  manifest, manifest_path,
  sep = "\t", quote = FALSE, row.names = FALSE, na = ""
)
print(manifest)
```

The manifest is a record of the review. Computing a digest does not by itself
review a biological assignment or grant distribution rights. If a table or
its registry metadata changes after review, review it again and assign a new
version. Do not replace the digest under an existing version.

Choose logical IDs in your own namespace. The `my_lab_...` IDs above are
tutorial examples and do not reuse a built-in lisaR identity.

## 6. Register and validate the same tables persistently

This block reads the manifest just written and installs those same three files
in lisaR's ordinary resource cache and registry. The registration remains
available to later R sessions. Nothing deletes the cache when the example
finishes.

```{r register-persistent, eval=FALSE}
manifest <- utils::read.delim(
  manifest_path, sep = "\t", quote = "", comment.char = "",
  check.names = FALSE, stringsAsFactors = FALSE,
  colClasses = "character", na.strings = character()
)

installed <- lapply(seq_len(nrow(manifest)), function(i) {
  row <- manifest[i, , drop = FALSE]
  install_lisa_resource(
    source = file.path(resource_dir, row$artifact),
    logical_id = row$logical_id,
    version = row$version,
    species = row$species,
    modality = row$modality,
    schema = row$schema,
    approved_origin = row$approved_origin,
    expected_sha256 = row$sha256,
    compatibility = row$compatibility
  )
})

refs <- setNames(
  paste0(manifest$logical_id, "@", manifest$version),
  manifest$schema
)

checked <- validate_lisa_custom_resources(
  dictionary_resource = refs[["lisa_dictionary@2"]],
  category_map_resource = refs[["category_map@1"]],
  term2gene_resource = refs[["term2gene@1"]],
  species = "Homo sapiens"
)

print(checked$resources[, c("config_key", "resource_id", "schema", "ready")])
print(as.data.frame(checked$summary, stringsAsFactors = FALSE))
stopifnot(checked$valid)
```

For this example, the summary should report 48 dictionary assignments, 4
categories, 48 gene sets, 576 TERM2GENE memberships and 40 genes. The
validator also checks the registered schema, species, profile compatibility,
digest and cross-table relationships. It cannot judge whether a biological
classification is appropriate.

Running the block again with the same identities, bytes and metadata reuses
the immutable registrations. Changed content or metadata requires a new
version.

## 7. Select the registered bundle in the sample study

Use the references created from that same manifest. Remove
`pipeline.lisa_dictionary`, because `core` and `expanded` select built-in
dictionaries.

```{r select-bundle, eval=FALSE}
study <- yaml::read_yaml(file.path(project, "study.yml"))
study$pipeline$workers <- 1L
study$pipeline$lisa_dictionary <- NULL
study$pipeline$dictionary_resource <- refs[["lisa_dictionary@2"]]
study$pipeline$category_map_resource <- refs[["category_map@1"]]
study$pipeline$term2gene_resource <- refs[["term2gene@1"]]

custom_config <- file.path(project, "study-custom.yml")
yaml::write_yaml(study, custom_config)

validation <- validate_lisa_config(custom_config, check_files = TRUE)
print(validation$resources[, c(
  "config_key", "configured_id", "resource_id", "species",
  "schema", "ready", "message"
)])
stopifnot(validation$execution_ready)
```

An execution-ready result means that the configuration, sample inputs and
registered resources passed the preflight checks. Validation does not run the
analysis.

## 8. Run, verify and recognise the custom output

After validation passes, run the same configuration and open its report:

```{r run-custom-study, eval=FALSE}
result <- run_lisa(custom_config)
check <- verify_lisa_run(result$output_dir)
stopifnot(identical(result$gate, "PASS"), identical(check$gate, "PASS"))

resolved <- utils::read.delim(
  file.path(result$output_dir, "config", "resource_resolution.tsv"),
  sep = "\t", quote = "", comment.char = "", check.names = FALSE
)
print(resolved[, c("config_key", "configured_id", "resource_id", "schema")])

browseURL(file.path(result$output_dir, "report_index.html"))
```

The resource table must show these exact selected identities:

* `my_lab_sample_dictionary@1.0.0`
* `my_lab_sample_category_map@1.0.0`
* `my_lab_sample_term2gene@1.0.0`

In the report, open **Synthetic response A** and then its **GOBP-C2** category
view. Custom results use the category labels **Example of signal response**,
**Example of stress response**, **Example of proliferation** and **Example of matrix remodelling**. The category map places the first two under **Example response programmes** and the others under **Example growth and structure**.

A category appears in a particular plot only when it has relevant member
gene-set results under that plot's filters. The exact resource identities in
`config/resource_resolution.tsv` and `resource_inventory.tsv` distinguish a
filtered category from accidental use of a built-in bundle.

In a completed configured study, an applicable custom semantic collection also
participates in the per-analysis category-inference family with the other
configured semantic collections. This is independent of how the dictionary
was constructed; it does not rebuild assignments. Check the exact collection
and category in the saved inference table before interpreting any stars. A
historical descriptive figure cannot acquire a category P from its mean NES.

See [Resources and provenance](resources-and-provenance.html) for registry and
licensing details, and [Configuration reference](configuration-reference.html)
for the remaining study settings.
