---
title: "lisaR function reference"
lang: en-GB
output:
  rmarkdown::html_vignette:
    mathjax: null
    includes:
      in_header: vignette-layout.inc
    toc: true
vignette: >
  %\VignetteIndexEntry{lisaR function reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use this page when you know the task but need the public R function, its main
arguments and the result to inspect. Examples that refer to your files or
objects are templates to adapt in an installed lisaR session. The small
DE-adapter example runs offline.

For a complete first analysis, use [Quick Start](quick-start.html). For a new
study, prepare the input, validate the configuration, inspect the plan, run the
study and verify the completed result.
You do not need a separate function to add standard category, gene or contrast
evidence: configured `run_lisa()` builds those products by default.

| What you want to do | Function | Writes files? |
|---|---|---|
| Prepare an already computed DE table, without fitting a model | `prepare_lisa_de_input()` | No |
| Copy the offline sample project | `lisa_init_project()` | Yes, to a new directory |
| Check whether a study is ready | `validate_lisa_config()` | No |
| Install a verified Riaz or CPTAC example input bundle | `install_lisa_example_bundle()` | Yes, to a new directory |
| Prepare the pinned MSigDB TERM2GENE resource | `prepare_lisa_msigdb_resource()` | Yes, to the managed resource cache and registry |
| Install one reviewed local scientific resource | `install_lisa_resource()` | Yes, to the managed resource cache and registry |
| Check a custom dictionary, category map and TERM2GENE together | `validate_lisa_custom_resources()` | No |
| Estimate report size | `plan_lisa_outputs()` | No |
| Run a complete study | `run_lisa()` | Yes |
| Run one DE analysis or one profile contrast | `run_lisa_de()` / `run_lisa_contrast()` | Yes |
| Add category inference to a verified earlier run | `add_lisa_category_inference()` | Yes, in a new directory |
| Refresh only saved category-inference presentation | `refresh_lisa_support_grades()` | Yes, in a new directory |
| Plan extra figures from a completed run | `plan_lisa_extension()` | No |
| Create extra figures from a completed run | `render_lisa_categories()` | Yes, in a new directory |
| Check a completed run | `verify_lisa_run()` | No |
| Open a completed run in the Shiny explorer | `explore_lisa_run()` | Yes, to an external exploration workspace |
| Build an advanced saved-result request or export | Exploration functions listed below | Depends on the explicit action |

The table lists the public interface. Do not build a workflow around internal
`lisaR:::` helpers. The function help pages give complete signatures; the
sections below explain when to choose each entry point and what to inspect.

## `prepare_lisa_de_input()`

Use this for data frames already produced by DESeq2, edgeR or limma. It fits
no model and filters no genes by adjusted P value. Keep the complete tested
table and declare the direction of the effect and ranking.

```{r reference-prepare-input, eval=TRUE}
library(lisaR)
source_de <- data.frame(
  gene = c("GENE_A", "GENE_B", "GENE_C", "GENE_D"),
  log2FoldChange = c(1, -0.5, 0.1, NA_real_),
  stat = c(3, -1, 0.2, NA_real_),
  pvalue = c(0.001, 0.3, 0.8, NA_real_),
  padj = c(0.01, 0.6, 0.95, NA_real_)
)
prepared <- prepare_lisa_de_input(
  source_de, method = "DESeq2",
  columns = list(symbol = "gene"),
  positive_direction = "Treated relative to control"
)
stopifnot(nrow(prepared$data) == nrow(source_de))
prepared$data
prepared$row_audit[, c("source_row", "rank_eligible")]
```

The non-significant rows and permitted missing statistics remain. The missing
rank is marked ineligible, not silently replaced by zero. `method` declares
column conventions; it does not verify which software produced a table.

| Returned object | Use |
|---|---|
| `data` | Canonical columns `symbol`, `log2FoldChange`, `rank_value`, `pvalue`, `padj` after the declared duplicate policy |
| `full_data` | Canonical version of all input rows before duplicate selection/remapping |
| `column_mapping` | Source-to-canonical column mapping |
| `row_audit` | Row identity and eligibility, including missing statistics |
| `duplicate_resolution` | Explicit decisions about duplicate rows |
| `receipt` | Method/direction declarations and processing choices |

DESeq2 Wald `stat` and limma `t` are recognised signed statistics. edgeR needs
an explicitly supplied signed rank; an unsigned multi-df F/LR statistic is
not given a direction automatically. DESeq2 LRT statistics likewise need an
appropriate separately prepared signed rank. `rank_direction` can be
`same_as_effect` (default) or `opposite_to_effect`; the latter negates ranks
once and checks comparable nonzero signs.

The default duplicate policy is `error`; DE statistics must not be aggregated.
`na_policy = "retain"` is the default, while `"error"` rejects missing
statistics. Infinite/malformed values and missing gene identifiers are always
errors. `matrix_scale` records a declaration only: it does not inspect,
transform or configure an expression matrix. See
[Preparing DE inputs](preparing-de-inputs.html) for method-specific examples.

## `lisa_init_project()`

This creates a new copy of the installed synthetic sample project, with
inputs and resources for an offline example.

```{r reference-api-template-1, eval=FALSE}
root <- if (.Platform$OS.type == "windows") "D:/lisa" else "~/lisa"
dir.create(path.expand(root), recursive = TRUE, showWarnings = FALSE)
project <- lisa_init_project(file.path(root, "sample"))
config <- file.path(project, "study.yml")
```

`path` must name a directory that does not exist. The function copies the
configuration and synthetic inputs, then returns the new project path
invisibly. It stops rather than merging with or overwriting an existing
directory. Use it to inspect a complete offline example before adapting a
configuration to your own data.

## `validate_lisa_config()`

Use this before every run. It checks a complete study definition without
running enrichment or creating analysis output.

```{r reference-api-template-2, eval=FALSE}
validation <- validate_lisa_config(
  config,
  check_files = TRUE,
  strict = TRUE
)
```

- `config`: YAML/JSON path or equivalent named R list.
- `check_files`: verify referenced input files and resolve the dictionary,
  TERM2GENE and category-map IDs against the active species, registry and
  managed cache. `FALSE` is structural prevalidation only and cannot establish
  execution readiness.
- `strict`: reject unknown keys at every configuration level. Use
  `strict = FALSE` only to inspect an older configuration before migrating it.
  Recognised fields, scientific contracts and security rules still apply.
  `run_lisa()` always validates strictly.

The result answers two questions: “is the structure valid?” and “can this
study run here?”. Its readiness stages are:

| Stage | Question answered |
|---|---|
| `structural_valid` | Does the configuration follow schema 1.0.0? |
| `inputs_ready` | Are the declared study inputs present and acceptable? |
| `dependencies_ready` | Are the required R packages available? |
| `resources_ready` | Do the registered scientific resources resolve correctly? |
| `execution_ready` | Have all required stages passed? |

The `readiness` table summarises those stages. The `inputs`, `dependencies`
and `resources` tables provide row-level diagnostics. `valid` is retained as
an alias for `execution_ready`, so a missing or unknown resource can never
return `valid = TRUE`. Structural contract errors raise an actionable error;
missing files, packages or registered resources return a summary with
`execution_ready = FALSE`. With `check_files = FALSE`, input/resource readiness
is not established, even when `structural_valid` is `TRUE`.

Inspect `validation$config` for the normalised configuration and
`validation$configuration_provenance` to distinguish supplied values from
defaults. Paths in YAML/JSON start at the study file's directory; paths in an
R list start at the current working directory. See the complete offline
[configuration example](configuration-reference.html).

## `install_lisa_resource()`

Use this after you have obtained and reviewed a local scientific resource. It
copies one immutable TSV into lisaR's managed cache and adds its versioned
identity to the local registry.

```{r reference-api-template-3, eval=FALSE}
installed <- install_lisa_resource(
  source = dictionary_path,
  logical_id = "lab_dictionary",
  version = "1.0.0",
  species = "Homo sapiens",
  modality = "transcriptomic/genomic",
  schema = "lisa_dictionary@2",
  approved_origin = "Local reviewed dictionary release 1.0.0",
  expected_sha256 = reviewed_dictionary_sha256
)
installed$resource_id
```

`source` must be an existing local, non-symbolic-link TSV. The SHA-256 must
come from the reviewed manifest for that resource; the function does not
download data, accept licence terms or turn an unknown file into an approved
resource. Install the dictionary, category map and TERM2GENE separately, using
the matching schema for each file.

`version = "1.0.0"` is the resource's first-publication version.
`schema = "lisa_dictionary@2"` is the score-free TSV contract, not a second
resource version. The legacy `lisa_dictionary@1` reader is only for an existing
historical resource; it is not the schema for a new score-free dictionary. The
package bundles the core and expanded LISA dictionaries and category map.
MSigDB TERM2GENE memberships remain external and must be acquired and
registered through an authorised route.

## `install_lisa_example_bundle()`

Use this to obtain one reviewed, version-pinned example input bundle before
running its public preparer. It supports `"riaz-gse91061"` and
`"cptac-ccrcc"`; it does not fit models or run LISA.

```{r reference-api-template-example-bundle, eval=FALSE}
bundle <- install_lisa_example_bundle(
  example = "riaz-gse91061",
  destination = "riaz-gse91061-bundle"
)
bundle$next_command
```

`destination` must be new. The optional `source` selects one local archive or
one explicit archive address, but cannot alter the package-pinned archive or
member identities. The returned `next_command` points to the example's
prepare-only route. Review and adapt the resulting project before a separate
scientific run.

## `prepare_lisa_msigdb_resource()`

Use this specialised helper for the default external
`msigdb_term2gene@2026.1` membership resource. It either transforms the exact
reviewed local MSigDB ZIP or, when no ZIP is supplied, acquires the one pinned
archive after the caller has reviewed the applicable terms.

```{r reference-api-template-msigdb, eval=FALSE}
term2gene <- prepare_lisa_msigdb_resource(
  source_zip = "msigdb.2026.1.zip"
)
term2gene$resource_id
```

For the acquisition route, call `prepare_lisa_msigdb_resource(accept_terms =
TRUE)` only after reviewing the applicable MSigDB, KEGG and BioCarta terms for
your intended use. That argument permits this call's acquisition only; it is
not a licence grant. If the exact resource is already registered, lisaR
resolves it without acquiring or transforming another copy.

## `validate_lisa_custom_resources()`

Use this once all three custom resources are registered. It checks their
schemas, species, immutable identities and cross-file relationships before a
study uses them.

```{r reference-api-template-4, eval=FALSE}
checked <- validate_lisa_custom_resources(
  dictionary_resource = "lab_dictionary@1.0.0",
  category_map_resource = "lab_category_map@1.0.0",
  term2gene_resource = "lab_term2gene@1.0.0",
  species = "Homo sapiens"
)
stopifnot(checked$valid)
```

See [Custom dictionaries and supercategories](custom-dictionaries-and-supercategories.html)
for the accepted TSV columns, relationships and a complete construction,
registration and selection example.

## `run_lisa()`

Use this for the normal configuration-defined workflow. With the default
`dry_run = TRUE`, it creates planning evidence only. After you deliberately set
`dry_run = FALSE`, it validates and runs every configured analysis and contrast,
then builds the requested report products.

```{r reference-api-template-5, eval=FALSE}
result <- run_lisa(config)
```

`config` is a YAML/JSON path or an equivalent named list. Relative paths in a
file configuration are resolved from the configuration file's directory;
relative paths in a list are resolved from the current working directory. The
run result is returned invisibly. With `dry_run = TRUE` (the default), it
contains planning evidence and `plan_dir`, not completed scientific results.
With `dry_run = FALSE`, the requested products are written below the new
`pipeline.output_dir` and verified before promotion.

The standard report includes the category navigator and evidence sheets,
global gene search and aligned evidence for declared compatible semantic
contrasts. A full `run_lisa()` also calculates a separate serial category-
inference stage across configured semantic collections within each analysis;
`HALLMARKS` and contrasts do not receive this inference. The default category
view shows adjusted-P stars up to `****` at P ≤0.0001, with exact P and d/N in
the category card. `report.category_nes_variants` defaults to all four of `clean`,
`percentages`, `direction` and `dispersion`; a non-empty unique subset is allowed. GeneCards
are opt-in (`report.legacy_gene_products = TRUE`), and ORA remains separate
and off by default (`pipeline.run_ora = FALSE`). These are configuration
fields, not extra `run_lisa()` arguments.

Parallelism is configured only through `pipeline.workers`; it is not a
separate `run_lisa()` argument. Independent tasks use stable identifiers and
fixed random seeds. lisaR waits for every worker to finish and checks the
result before placing it in the final output folder. See
[Parallel execution](parallel-execution.html).

At the start, lisaR binds the study to one SHA-256 contract. That contract
covers the exact configuration, every original scientific input, the three
registered resources, the installed lisaR code and the R/package dependency
closure used for this run.

| Artefact | What it records |
|---|---|
| `run_contract.tsv` | The canonical digest that binds all contract components |
| `input_inventory.tsv` | DE, matrix, mapped-ID, external-index, allowlisted source-data and duplicate-mapping inputs |
| `resource_inventory.tsv` | The three registered scientific resources and their identities |
| `executable_code_inventory.tsv` | The installed lisaR code used by the run |
| `effective_environment.tsv` | The R runtime and installed dependency closure that actually ran |

These receipts use roles, portable identifiers or base file names and SHA-256
values, not absolute source paths. `effective_environment.tsv` is provenance;
it is not an environment lock or an installer.

## Reuse verified saved results for category inference

`add_lisa_category_inference(source_run, output_dir, term2gene = NULL)` adds
inference to a **new copy** of an earlier complete run using its verified saved
fgsea results and exact TERM2GENE identity. It does not rerun DE or GSEA. Use
this only when inference is genuinely absent; it calculates new category P
values and simultaneous lower bounds from the saved gene-set results. The
source remains unchanged. See `?add_lisa_category_inference` for the full
input contract and plot formats.

`refresh_lisa_support_grades(source_run, output_dir)` instead makes a new,
verified copy of a run that **already has category inference**. It refreshes
the current four-star presentation and report sources without recalculating
DE, GSEA, closed tests, category P, d/N or hypothesis families. The destination
must not exist, and a new report must be built and sealed in that copy. This is
the correct route for changing star presentation on an accepted inferred run;
it is not a way to add inference to `run_lisa_de()` output by assumption.

## `plan_lisa_outputs()`

Use this before a large or full report when you need to judge storage and file
counts. It estimates report size without starting enrichment.

```{r reference-api-template-6, eval=FALSE}
plan <- plan_lisa_outputs(config)
plan$summary
plan$products
```

`config` is a YAML/JSON path or the equivalent named list. If known, pass a
named `category_counts` vector to replace the conservative category-count
assumption, and use `bytes_per_figure` to adjust the storage estimate. The
function returns summary, collection, product and file-policy tables. It does
not run enrichment, resolve a missing resource into existence or establish
execution readiness. Unspecified category counts use a conservative 100 per
collection; the default storage assumption is 1 MiB per figure format.
The estimate includes standard evidence/member figures and selected NES
variants, but gene-search payload size depends on complete memberships and DE
rows. Treat this as an estimate, not a memory or disk-space guarantee.

## `run_lisa_de()`

Use this lower-level function for one LISA analysis of a prepared DE result
outside a configuration-defined study. For routine work, prefer `run_lisa()` so
the settings are stored with the result.

The available collection IDs are `GOBP-C2`, `GOMF`, `GOCC`, `PATHWAYS` and
`HALLMARKS`. If `universes` is omitted, lisaR uses all five. The example below
selects the four semantic LISA collections; `HALLMARKS` remains available as a
direct MSigDB analysis but is deliberately not part of this example.

```{r reference-api-template-7, eval=FALSE}
run_lisa_de(
  input,
  output_dir,
  comparison_name = NULL,
  species = "Homo sapiens",
  symbol_col = NULL,
  rank_col = NULL,
  logfc_col = NULL,
  pvalue_col = NULL,
  padj_col = NULL,
  universes = c("GOBP-C2", "GOMF", "GOCC", "PATHWAYS"),
  lisa_dictionary = "core",
  term2gene_path = NULL,
  make_heatmaps = FALSE,
  ...,
  trusted_rds = FALSE,
  rds_max_bytes = 512 * 1024^2
)
```

If `input` is an RDS file, trust is the first decision. Set
`trusted_rds = TRUE` only for a local, stable file under your control and from
a source you fully trust. The `rds_max_bytes` checks are useful diagnostics,
but they are not a sandbox and cannot make an untrusted RDS safe.

Arguments:

- `input`: DE table path, data frame or supported DE result object.
- `output_dir`: destination directory.
- `comparison_name`: stable portable ASCII technical identifier used in paths
  and file names. Put Unicode or spaced display text in a separate title or
  label field.
- `species`: `Homo sapiens` or `Mus musculus`.
- `symbol_col`, `rank_col`, `logfc_col`, `pvalue_col`, `padj_col`: explicit
  source columns. `rank_col` is optional when a compatible effect size exists.
- `universes`: one or more of the five collection IDs listed above.
- `lisa_dictionary`: `core` by default, resolving the registered
  `lisa_dictionary_core@1.0.0` resource; `expanded` resolves
  `lisa_dictionary_expanded@1.0.0`. Both score-free dictionary resources, and the
  category map, are bundled with lisaR. The default MSigDB TERM2GENE
  memberships are external. If all three explicit resource paths are passed
  through `...`, this argument is a portable label for the custom dictionary
  instead of a registered tier.
- `term2gene_path`: optional path to a species-matched gene-set membership
  table. Omit all three resource paths to use registered defaults, or provide
  `term2gene_path`, `lisa_dictionary_path` and `category_map_path` together for
  an advanced explicit-path workflow.
- `make_heatmaps`: reserved compatibility flag. In version 1.0 the lower-level
  function warns and does not generate heatmaps. For heatmaps, use a configured
  `run_lisa()` study with an expression matrix, then request the `heatmap`
  product through `render_lisa_categories()`.
- `...`: advanced single-analysis arguments. Every argument passed through
  `...` must be named explicitly; unnamed advanced arguments are rejected
  before the analysis engine is called.
- `trusted_rds`: named-only confirmation that the caller accepts the risks of
  reading a **local, stable and fully trusted** `.rds` file. YAML/JSON study
  configurations cannot grant this capability and should use TSV/CSV inputs.
  Supported serialised DE inputs are
  tables/matrices, `DESeqDataSet`, `DESeqResults`, `DGELRT`, `DGEExact` and
  `TopTags`; `DGEList` is rejected because it contains counts rather than a DE
  result. When gene identifiers are stored in row names, use
  `symbol_col = "rowname"` explicitly.
- `rds_max_bytes`: named-only positive whole-byte diagnostic limit applied
  before deserialisation and again to the approximate in-memory object size.
  It is not a sandbox or a hard memory bound against hostile or highly
  compressed RDS content, symlink changes or file replacement between checks
  and deserialisation. `trusted_rds = TRUE` therefore means that the caller
  accepts those risks for a stable file under their control.

### DESeq2, edgeR and trusted RDS inputs

Objects already loaded in the current R session can be passed directly:

- `DESeqResults` is used as the DE result table.
- `DESeqDataSet` requires exactly one selector: `deseq2_name` or
  `deseq2_contrast`, supplied as a named argument through `...`.
- edgeR `DGELRT`, `DGEExact` and `TopTags` results are accepted.
- `DGEList` is rejected because it contains counts rather than a fitted DE
  result.

For example, this selects one DESeq2 coefficient and the four semantic
collections:

```{r reference-api-template-8, eval=FALSE}
run_lisa_de(
  input = dds,
  output_dir = "results/treated-vs-control",
  comparison_name = "treated_vs_control",
  species = "Homo sapiens",
  symbol_col = "rowname",
  universes = c("GOBP-C2", "GOMF", "GOCC", "PATHWAYS"),
  deseq2_name = "condition_treated_vs_control"
)
```

Use `deseq2_contrast = c("condition", "treated", "control")` instead when a
three-part contrast is the intended DESeq2 result; do not provide both
selectors. A local `.rds` path additionally requires `trusted_rds = TRUE`.
YAML and JSON configurations accept tabular inputs only and can never grant
permission to read an RDS file. Export a TSV/CSV when the configured workflow
is required.

Each accepted RDS input produces
`qc/<comparison>_trusted_RDS_input_receipt.tsv` with schema
`lisa_rds_input_receipt_v2`.

| Receipt field | Meaning |
|---|---|
| `source_sha256` | SHA-256 of the serialised source file |
| `object_sha256` | SHA-256 of the validated in-memory object serialised with R version 3/XDR |
| File and object details | Base file name, class, dimensions, serialised size and approximate in-memory size |
| Trust and limit details | Configured diagnostic limit and explicit authorisation state |

`object_sha256` protects integrity while the object passes through internal
branches, but it is not a stable semantic identity across R or package
versions. Absolute source paths are not written to the receipt. The same
receipt is retained when one call branches over several LISA collections or
dictionaries, and the RDS is deserialised only once.

Advanced arguments accepted through `...` include input type, MSigDB mode,
dictionary/category-map paths, C2 sources, DESeq2 contrast/name, GSEA and
over-representation analysis (ORA)
cutoffs, gene-set-size limits, thread count, random seed, output/plot formats,
plot ordering, palette, overwrite and verbosity. These are lower-level options;
configuration-defined studies are preferred because they save the settings
used for the analysis beside the results.

## `run_lisa_contrast()`

Use this lower-level function to compare two compatible, completed
single-analysis results. Prefer declaring the contrast in `run_lisa()` when it
belongs to a complete study.

```{r reference-api-template-9, eval=FALSE}
run_lisa_contrast(
  contrast_a,
  contrast_b,
  output_dir,
  comparison_name = NULL,
  contrast_a_label = "Contrast A",
  contrast_b_label = "Contrast B",
  universes = c("GOBP-C2", "GOMF", "GOCC", "PATHWAYS"),
  ...
)
```

- `contrast_a`, `contrast_b`: compatible completed single-analysis outputs.
- `output_dir`: contrast destination.
- `comparison_name`: stable contrast identifier.
- `contrast_a_label`, `contrast_b_label`: labels used in tables and plots.
- `universes`: collections to compare. The five available IDs are
  `GOBP-C2`, `GOMF`, `GOCC`, `PATHWAYS` and `HALLMARKS`; omitting the argument
  uses all five. The example selects the four semantic collections and leaves
  out direct `HALLMARKS` analysis deliberately.

The low-level call computes descriptive profile comparisons. It is not the
complete report/evidence pipeline. Declare `contrasts` in a configured study
to obtain the linked A/B navigator, aligned matrices and return links to genes.
The two NES distributions and significance states remain separate; a delta
is not a category-level significance test. Legacy inputs without member
provenance cannot supply fabricated medians or percentiles.

In a configured study, `pipeline.run_hallmarks: false` removes `HALLMARKS`
from the effective collection selection. `true` permits it, but does not add
it to an explicit `collections` list that omits it.

Advanced options through `...` control plot subsets/order, missing categories,
direction tolerance, grouping, exports, plot formats/background, gene-set
annotation, file labels, title prefix, overwrite and verbosity. Prefer
declaring contrasts in `run_lisa()` unless constructing an advanced workflow.

## `plan_lisa_extension()`

Use this to check which extra figures would be created from an existing
standard run. Planning is read-only.

```{r reference-api-template-10, eval=FALSE}
selection <- list(
  analyses = "analysis_a",
  categories = c("SYN_CATEGORY_A", "SYN_CATEGORY_B"),
  products = c("gene_cards", "volcano")
)
plan <- plan_lisa_extension("results/study", selection)
plan[c("mode", "expected_files", "selected_categories")]
```

`source_run` must be a completed run that passes `verify_lisa_run()`.
`selection` may be `"full"`, a selection-file path or a named list. Planning
is read-only and stops if a selection name is invalid or the source run has
changed.

## `render_lisa_categories()`

Use this after reviewing an extension plan. It creates additional figures from
a completed standard run without repeating enrichment.

```{r reference-api-template-11, eval=FALSE}
rendered <- render_lisa_categories(
  "results/study",
  selection,
  output_dir = "results/study-selected"
)
```

`output_dir` must be a new directory outside the source run. The function
reuses the saved enrichment tables, creates and checks the requested figures,
and returns a receipt invisibly. It does not repeat enrichment or alter the
source run. It stops if a requested product cannot be built or validated.

## `verify_lisa_run()`

Use this before accepting, sharing or interpreting a run. It checks that the
expected files exist, are unchanged and passed lisaR's final validation.

```{r reference-api-template-12, eval=FALSE}
check <- verify_lisa_run("results/example")
check$gate
check$findings
```

`run_dir` is the completed results folder. The return value contains `gate`
(`PASS` or `FAIL`), detailed findings and the path to the file list used for
verification. A PASS means that every expected file is present and unchanged
and that lisaR finished its final checks. It does **not** establish biological
validity, a valid model design, redistribution rights or pixel-identical figure
reproduction on other systems. The [troubleshooting guide](troubleshooting.html)
explains those distinctions and how to preserve a failed run for diagnosis.

The main verification artefacts are:

| Artefact | Role in verification |
|---|---|
| `expected_artifacts.tsv` | States which products are required, optional or not applicable |
| `artifact_validation.tsv` | Reconciles required products with their status and material files |
| `run_manifest.tsv` | Lists final files, sizes and SHA-256 values |
| `run_events.tsv` | Records that final validation and promotion completed |

## Explore or export a saved result

Use `explore_lisa_run()` for the supported interactive route. It requires the
optional Shiny package, verifies the completed run and opens it without
changing the source result:

```{r reference-explore-saved-run, eval=FALSE}
explore_lisa_run("results/example")
```

The exploration workspace is outside the immutable source run. Browsing reads
saved material. A selected figure is rendered only after an explicit request,
and this rendering does not repeat DE, GSEA or ORA. Follow
[Explore saved results with Shiny](shiny-app.html) for the normal interactive
workflow and portable export.

The lower-level exploration API is also public for advanced interfaces:

| Task | Public functions |
|---|---|
| Open a workspace and inspect available figures | `lisa_explore_open()`, `lisa_explore_catalog()`, `lisa_explore_status()` |
| Define and inspect one exact request | `lisa_figure_request()`, `lisa_request_id()`, `lisa_explore_plan()` |
| Submit and follow explicit work | `lisa_explore_submit()`, `lisa_explore_poll()`, `lisa_explore_cancel()` |
| Inspect reusable outputs | `lisa_explore_artifacts()`, `lisa_explore_stale_artifacts()` |
| Export validated available material | `lisa_explore_export()` |

`lisa_explore_submit()` schedules one exact figure. It is the boundary between
read-only navigation and rendering. `lisa_explore_export()` accepts `dir` or
`zip` output and snapshots only validated material already available when the
export begins. It does not generate missing products. Native KEGG and contrast
preparation use the additional exported helpers
`lisa_explore_configure_kegg()`, `lisa_explore_prepare_kegg_index()` and
`lisa_explore_prepare_contrast_products()`. Use their installed help pages
because they require an existing local snapshot or compatible saved evidence.
