---
title: "Run lisaR with your prepared files"
lang: en-GB
output:
  rmarkdown::html_vignette:
    mathjax: null
    includes:
      in_header: vignette-layout.inc
vignette: >
  %\VignetteIndexEntry{Run lisaR with your prepared files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

For the short first-run route, use [Get started](documentation.html). This
guide keeps the extended R examples and configuration options.

Use this guide to run lisaR on your differential-expression results and open
the HTML report. You do not need to write a configuration file or choose the
LISA dictionaries. The default output includes PNG and SVG figures, tables
and R scripts for reproducing the figures.

Install lisaR first, then follow the steps below. If you already use a YAML or
JSON study file, you can continue to pass it to `run_lisa()`.

If you want to practise without study data, use the self-contained
[Quick Start](quick-start.html). Its files and resources are synthetic and must
not be reused as the gene universe for a real study.

## What you need

- a complete differential-expression table, not only the significant rows;
- one sentence stating what a positive effect means in your comparison; and
- gene-set memberships for your species, prepared in the first step below.

[Prepare differential-expression inputs](preparing-de-inputs.html) explains the
table conventions. DESeq2 tables use the usual column names by default.
For edgeR or limma tables, specify `method = "edgeR"` or `method = "limma"`.
The ranking statistic must have a sign; the optional input examples below
show how to declare it.

## 1. Prepare the gene-set membership resource, once

lisaR ships its LISA dictionaries and category hierarchy. They assign gene sets
to biological categories, but they do not contain the genes in each set. That
membership comes from MSigDB and is prepared once per machine.

Read the [MSigDB terms](https://www.gsea-msigdb.org/gsea/msigdb_license_terms.jsp)
before running this command. Set `accept_terms = TRUE` only if you accept them.

```{r prepare-membership, eval=FALSE}
library(lisaR)

msigdb <- prepare_lisa_msigdb_resource(accept_terms = TRUE)
```

If you already hold an authorised MSigDB ZIP, use the local-file form described
in [Resources and provenance](resources-and-provenance.html). Do not point a
real study at the sample project's 40-gene TERM2GENE resource.

After registration, `run_lisa()` selects the resources for the species you
specify. If a required resource is missing, it tells you what to prepare
before starting the analysis.

## 2. Run one analysis

This is a complete first run. Replace the path, the species and the direction
sentence with your own.

```{r run-one, eval=FALSE}
result <- run_lisa(
  de                 = "de_treated_vs_control.tsv",
  output_dir         = "lisa-treated-vs-control",
  species            = "Homo sapiens",
  positive_direction = "higher in treated than in control"
)

```

`de` also accepts the data frame itself or a DESeq2 Wald-test
`DESeqResults` object. For DESeq2 likelihood-ratio tests and edgeR results,
see the input examples below. `positive_direction` has no default: lisaR never infers which
side of a comparison a positive effect belongs to.

Choose a new folder for `output_dir`; lisaR will create it. If an input or
argument is rejected before the analysis starts, correct it and repeat the
call with the same folder name.

By default this produces a standard report with PNG and SVG figures, the source
data tables and the figure reproduction recipes. PDF is off unless you ask for
it.

The configured `run_lisa()` workflow calculates category inference after the
gene-set analyses. For each semantic collection, the default category view
shows adjusted-P stars (up to `****` at P ≤0.0001); the card retains the exact
P and minimum supported count d/N. This is separate from the descriptive
mean NES and from a comparison between two completed analyses.

Check that the analysis completed successfully:

```{r verify-one, eval=FALSE}
stopifnot(identical(result$gate, "PASS"))

check <- verify_lisa_run(result$output_dir)
stopifnot(identical(check$gate, "PASS"))
```

These checks should both return `"PASS"`. Verification checks the saved
files without repeating the analysis. You can then open the report:

```{r open-one, eval=FALSE}
browseURL(file.path(result$output_dir, "report_index.html"))
```

`result$reports` lists the available report pages. Keep the whole output
folder together when moving or sharing it so that the links continue to work.

If either gate is not `"PASS"`, keep the message and the output folder, and
follow [Troubleshooting](troubleshooting.html).

## 3. Compare two analyses

A contrast compares two completed LISA profiles. Declare the orientation
explicitly: `contrast = c("ON", "PRE")` means **ON minus PRE**. The order of
`de` never decides it.

```{r run-two, eval=FALSE}
result <- run_lisa(
  de = list(
    PRE = "de_pre_vs_control.tsv",
    ON  = "de_on_vs_control.tsv"
  ),
  output_dir = "lisa-pre-on",
  species    = "Homo sapiens",
  positive_direction = list(
    PRE = "higher at baseline than in its control",
    ON  = "higher on therapy than in its control"
  ),
  contrast = c("ON", "PRE"),
  title    = "Baseline versus on-therapy"
)

result$reports[result$reports$page == "contrasts", ]
```

The contrast page states the declared orientation. A positive value means a
larger mean NES in `ON` than in `PRE`. That mean is descriptive: it is not a
category-level significance test, and it is not another
differential-expression test.

For more than one contrast, supply a list of pairs. Each pair specifies its
own order.

## Optional settings

The steps above are enough to run an analysis and read its report. The
sections below cover output options and less common input formats.

### Adjust what is produced

The first example needs no output settings. Add options only when you want
to change what lisaR produces:

```{r adjust, eval=FALSE}
result <- run_lisa(
  de                 = "de_treated_vs_control.tsv",
  output_dir         = "lisa-study",
  species            = "Homo sapiens",
  positive_direction = "higher in treated than in control",
  svg                = FALSE,       # PNG only
  recipes            = FALSE,       # no figure recipes
  collections        = c("GOBP-C2", "GOMF"),
  workers            = 4L
)
```

Use `dry_run = TRUE` to inspect a plan without calculating enrichment or
building the report.

For a new analysis, `mode = "full"` writes one report in the same
`output_dir`: the standard report plus every applicable FULL figure (member
gene sets, prioritized genes, volcano overlays, gene heatmaps, recurrent
genes, contrast profiles and paired products), each placed in its analysis,
collection and category. FULL can take substantially longer and use several
gigabytes. To add figures to an existing STANDARD result without repeating
enrichment, use `plan_lisa_extension()` followed by `render_lisa_categories()`,
which writes its own gallery, as shown in
[The HTML report](scientific-report.html).

For settings without a dedicated argument, use `pipeline` or `report`.
Unspecified settings keep their defaults:

```{r advanced-overrides, eval=FALSE}
result <- run_lisa(
  de                 = "de_treated_vs_control.tsv",
  output_dir         = "lisa-study",
  species            = "Homo sapiens",
  positive_direction = "higher in treated than in control",
  report             = list(evidence_max_sets = 40L),
  pipeline           = list(run_hallmarks = TRUE)
)
```

Specify each setting in one place: as a direct argument, or inside `pipeline`
or `report`. Repeating a setting is rejected, even when both values agree.

### Tables that need an explicit column mapping

lisaR reads the usual DESeq2, edgeR and limma column names. Name the columns
when yours differ, or when the statistic to rank on is not the conventional
one:

```{r columns, eval=FALSE}
result <- run_lisa(
  de                 = "limma_treated_vs_control.tsv",
  output_dir         = "lisa-limma-study",
  species            = "Homo sapiens",
  positive_direction = "higher in treated than in control",
  method             = "limma",
  columns            = list(symbol = "gene", rank = "t")
)
```

Some ranking statistics have no sign and need an explicit choice:

- a DESeq2 likelihood-ratio result, whose `stat` is unsigned;
- an edgeR result, whose `F` and `LR` statistics are unsigned.

For an edgeR **one-dimensional comparison**, one possible ranking is the
signed square root of its quasi-likelihood F statistic. This example is not
valid for a joint test of several coefficients:

```{r edger, eval=FALSE}
# `qlf` tests one contrast in your own edgeR model.
top <- as.data.frame(edgeR::topTags(qlf, n = Inf, sort.by = "none"))
top$symbol      <- rownames(top)
top$signed_stat <- sign(top$logFC) * sqrt(pmax(top$F, 0))

result <- run_lisa(
  de                 = top,
  output_dir         = "lisa-edger-study",
  species            = "Homo sapiens",
  positive_direction = "higher in treated than in control",
  method             = "edgeR",
  columns            = list(symbol = "symbol", rank = "signed_stat")
)
```

The signed statistic is your modelling decision; lisaR records it and does not
derive one for you.

`topTags()` supplies the adjusted p-values (`FDR`; BH adjustment by default).
If you pass a `DGEExact` or `DGELRT` object directly, its table must already
contain your adjusted p-values and signed ranking; name those columns with
`columns = list(padj = "your_q_value", rank = "your_signed_stat")`.
lisaR does not choose or recalculate the adjustment.

For a DESeq2 likelihood-ratio result, do not use its unsigned `stat` as the
ranking. Add a separately computed signed statistic for the comparison you
intend to interpret, then select it with `columns = list(rank = "signed_stat")`.

## Keep your results

Keep the complete study folder, including `inputs/`, `results/` and
`study.json`. The JSON file records the settings used and can be passed back
to `run_lisa()` with its full path.

Keep the whole folder together with your original export and its model record.
lisaR does not fit or refit the differential-expression model.

If a run fails after creating `study.json`, keep the error message and folder.
Consult [Troubleshooting](troubleshooting.html) before retrying the saved
configuration.

### Use a configuration file

Use a study file when you want to save and edit the complete configuration.
You can start from the `study.json` created by `run_lisa()`; the
[configuration reference](configuration-reference.html) explains its fields.

```{r config-route, eval=FALSE}
validation <- validate_lisa_config("study.yml", check_files = TRUE)
validation$readiness
stopifnot(validation$execution_ready)

result <- run_lisa("study.yml")
```

Existing configuration files keep their previous defaults. To include SVG and
recipes, use this structure:

```yaml
report:
  formats:
    svg: true
  recipes: true
```

When passing a configuration file, edit settings in that file rather than
adding the direct-input arguments to the same call.

## Continue with a prepared scientific example

The [Riaz RNA-seq example](riaz-gse91061-worked-example.html) and
[CPTAC proteomics example](cptac-ccrcc-proteomics.html) provide prepared DE
tables, source context and complete study configurations. Follow the instructions for the chosen example. Its gene-set memberships
belong to that example and should not be used for unrelated studies.
