---
title: "Licence, attribution and citation"
lang: en-GB
output:
  rmarkdown::html_vignette:
    mathjax: null
    includes:
      in_header: vignette-layout.inc
    toc: true
vignette: >
  %\VignetteIndexEntry{Licence, attribution and citation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Before sharing lisaR, a resource or a report, identify which material you are
sharing. The software licence, fixed LISA annotations, third-party scientific
resources and your own study material have different notices and conditions.
One licence does not replace the others.

Where a report presents category-level adjusted P and minimum supported d/N,
its methods use robust Hommel closed testing. Cite Goeman et al. (2019),
*Biometrika* 106, 841–856,
[doi:10.1093/biomet/asz041](https://doi.org/10.1093/biomet/asz041), alongside
the study's data and gene-set resource sources. This method attribution does
not replace provider licences or justify biological claims beyond the saved
analysis.

## Identify the material

| Material | Where its conditions come from | What to keep with it |
|---|---|---|
| lisaR code, documentation and synthetic fixtures | `DESCRIPTION` and `inst/COPYING` | Package name and version, GPL-3.0-or-later notice, copyright holder, author and corresponding source obligations when distributing covered copies. |
| Fixed LISA dictionaries and category map in this candidate | Installed resource manifest, provenance and third-party notices | Exact resource IDs, SHA-256 records, LISA/CSIC notice and applicable MSigDB, KEGG or BioCarta attribution. |
| External TERM2GENE memberships and native pathway material | The provider's current terms and the recipient's authorised acquisition route | Provider attribution, exact release or snapshot identity and any collection-specific notices. |
| R package dependencies | Each dependency's own licence | The notices and sources required for the form in which you distribute them. |
| Your input data and generated report | Your data source, consent/governance, study agreements and incorporated resources | Study citations, data provenance, resource inventory and any provider notices that apply to included material. |

This guide reports the notices installed with lisaR. It does not create a new
permission or settle whether a particular dataset or report may be distributed.

## Share the lisaR software under its licence

`DESCRIPTION` records `License: GPL (>= 3)`, corresponding to
GPL-3.0-or-later. The copyright holder is the Agencia Estatal Consejo Superior
de Investigaciones Científicas (CSIC), and the author is David Olmeda Casadomé.
The complete GPL text is installed as `COPYING`.

The licence covers lisaR code, documentation and original synthetic fixtures.
When you distribute a covered copy or modification, retain the licence and
copyright notice and meet the corresponding-source obligations for that form
of distribution. The GPL does not grant rights in external scientific
resources.

Source and issue access are identified in the package metadata:
[DBM-OlmedaLab/lisaR](https://github.com/DBM-OlmedaLab/lisaR).

## Retain the fixed-annotation notices

lisaR includes the exact
`lisa_dictionary_core@1.0.0`,
`lisa_dictionary_expanded@1.0.0` and
`lisa_category_map@1.0.0` resources. They record fixed
gene-set-to-category assignments and category organisation, not external gene
memberships.

The installed provenance records the CSIC/LISA annotations and retains
MSigDB, KEGG and BioCarta attribution where their identifiers occur.
The resource manifest identifies the bundled annotations and their
source-specific conditions. Do not treat `GPL-3.0-or-later` as permission to
redistribute third-party memberships or diagrams.

The installed third-party notice records that MSigDB identifies KEGG MEDICUS
content as CC BY-SA 4.0 and describes one-way compatibility from CC BY-SA 4.0
to GPLv3 for adapted material. That compatibility does not remove MSigDB or
KEGG attribution and does not replace collection-specific conditions.

The current technical reconstruction is tied to MSigDB 2026.1 through
`msigdbr` 26.1.0. The original historical model-assisted annotation retrieval
release is unrecorded. These are separate provenance facts. See
[How LISA dictionaries were built](dictionary-construction.html) for the
source-backed history and its limits.

## Keep external resources under their own terms

| External material | Installed notice |
|---|---|
| MSigDB identifiers or TERM2GENE memberships | Retain Molecular Signatures Database attribution and the applicable collection-specific conditions. Review the [MSigDB licence terms](https://www.gsea-msigdb.org/gsea/msigdb_license_terms.jsp). |
| KEGG MEDICUS or KEGG LEGACY identifiers, KGML or pathway images | Retain MSigDB and KEGG attribution where applicable and review the [KEGG copyright and licence information](https://www.kegg.jp/kegg/legal.html). A user-authorised cache snapshot is not part of the lisaR software licence. |
| BioCarta identifiers or memberships | Retain MSigDB and BioCarta acknowledgement and disclaimer conditions linked from the MSigDB terms. |

The MSigDB preparation helper asks a recipient to assert acceptance only before
the supported public acquisition. That assertion permits the call; it does not
grant a licence, accept terms for another person or confer redistribution
rights. A recorded resource ID or SHA-256 likewise proves identity, not
permission.

Dependencies retain their own licences. If you redistribute an installed
library, container or other combined form, review the actual dependency set and
the notices required for that distribution. Installing a dependency does not
replace the terms for scientific content retrieved through it.

## Share a report without changing its rights

Before sending or publishing a lisaR report:

1. Keep the report's configuration, input provenance,
   `resource_inventory.tsv`, `config/resource_resolution.tsv` and manifest.
2. Identify the lisaR package version and the source commit or archive used.
3. Retain the study and dataset citations required by their sources.
4. Retain the MSigDB, KEGG, BioCarta or other provider attribution that applies
   to memberships, identifiers, maps or derived figures in the report.
5. Check whether your agreements permit you to distribute the underlying data,
   external resources and generated material to the intended recipients.

lisaR does not place a new licence on a user's input data merely by processing
it. Conversely, generating an HTML page or figure does not remove conditions
attached to incorporated data or third-party resources.

## Cite the installed software and scientific sources

Inspect the citation generated from the exact installed package metadata:

```{r cite-lisar, eval=FALSE}
citation("lisaR")
packageDescription("lisaR")[
  c("Package", "Version", "Date", "URL", "Author")
]
```

Cite the version used in your analysis. For this release, R derives the software
citation from the installed package metadata.

Cite the study data and analytical methods used in your work. For external
gene-set resources, follow the provider's current citation instructions and
record the exact release used. The installed resource provenance names
Molecular Signatures Database, Liberzon et al. and Subramanian et al.; consult
the provider material for the complete citations appropriate to your selected
release and collections.

## Locate the exact installed notices

Use the installed paths rather than copying notices from another lisaR version:

```{r locate-licence-notices, eval=FALSE}
system.file("COPYING", package = "lisaR")
system.file("THIRD_PARTY_NOTICES.md", package = "lisaR")
system.file(
  "extdata", "dictionaries", "RESOURCE_PROVENANCE.tsv",
  package = "lisaR"
)
```

Read those files before redistribution. If the status or terms of material you
intend to share remain ambiguous, keep that material private until the relevant
rights holder or institutional process resolves the question.
