---
title: "Parallel execution"
lang: en-GB
output:
  rmarkdown::html_vignette:
    mathjax: null
    includes:
      in_header: vignette-layout.inc
    toc: true
vignette: >
  %\VignetteIndexEntry{Parallel execution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use this guide to choose a worker request, confirm what the platform actually
used and prepare a clean retry after a failure. `pipeline.workers` controls
concurrent independent tasks in a configuration-defined study. It is a
resource request, not a speed or memory guarantee. See
[Configuration reference](configuration-reference.html) for the complete field
contract and [Troubleshooting](troubleshooting.html) when a run stops.

## Choose workers and check the actual backend

The configuration schema default is `workers: 4`. Quick Start deliberately
overrides this with one worker for its demonstration run; that does not change the
schema default. On Linux, lisaR can use its multicore
backend for independent tasks. On Windows, macOS and other platforms it uses a
serial backend even when a larger number is requested. The effective worker
count is limited by task count, visible cores, recognised scheduler allocation
and R-check limits.

```yaml
pipeline:
  workers: 1
```

Use `workers: 1` for a serial route. Request more workers only within the CPU
and memory allocation of the current machine or scheduler job. lisaR does not
measure per-task memory or reserve RAM. Shared report assembly, final
verification and some cache operations remain serial.

The post-LISA category-inference stage is also serial across the completed
semantic collections of one analysis. `pipeline.workers` does not create
separate Hommel families or make category P values comparable as an A/B test.

| Situation | Current behaviour |
|---|---|
| Linux and more than one effective worker | Independent tasks may use the multicore backend. |
| Windows, macOS or other platform | Serial backend. |
| Fewer tasks than requested workers | Effective count is reduced to task count. |
| Scheduler or visible-core limit | Effective count is capped and recorded. |
| Insufficient memory or a worker error | The run stops; no partial result is promoted as complete. |

The supported scheduler limits include `SLURM_CPUS_PER_TASK`, `PBS_NP`,
`NSLOTS`, `NCPUS` and `LSB_DJOB_NUMPROC`. The record, rather than the request
alone, identifies the effective backend and worker count.

## Use one worker for the recorded contrast-evidence path

A Quick Start run with two DE analyses, one contrast, four semantic
collections and enabled category evidence failed with `workers: 4` while concurrently creating a contrast-evidence directory. The
recorded condition began:

```text
Could not create the destination parent ... report_pages/contrast_evidence/...
```

The same scientific scope completed with `workers: 1`, and
`verify_lisa_run()` returned `PASS`. Until this directory-creation issue is
resolved, use `workers: 1` for this contrast-evidence path. Other parallel
workloads were not covered by that check.

## Compare seeds, order and execution records

lisaR derives a task seed from a root seed and stable task identifier, records
the planned task context, and sets one inner numerical thread for each outer
task. These controls support reproducible task setup, but do not by themselves
guarantee identical numerical output for every operating system, package
version, library or workload.

After a completed run, inspect the execution receipt and task seeds:

```{r parallel-record-template, eval=FALSE}
run_dir <- "results/study"
execution <- read.delim(file.path(run_dir, "parallel_execution.tsv"))
seeds <- read.delim(file.path(run_dir, "parallel_task_seeds.tsv"))
execution
head(seeds)
```

`parallel_execution.tsv` records requested and effective workers, backend,
platform and `cap_reason`. `parallel_task_seeds.tsv` records task IDs and
seeds. Keep these with the configuration and resource provenance when
comparing runs.

## Keep cache controls separate from workers

The optional content cache stores reusable raw GSEA results. It is not a
resource registry and does not replace a completed run. Active cache modes
require an explicit local `pipeline.cache_dir`.

| `pipeline.cache_mode` | On a cache hit | On a miss |
|---|---|---|
| `off` | Does not read a cache | Computes without cache access. |
| `readonly` | Reuses the compatible entry | Computes without writing an entry. |
| `readwrite` | Reuses the compatible entry | Computes and attempts to store an entry. |
| `refresh` | Recomputes | Attempts to replace the entry. |

A hit requires the exact relevant ranking, selected memberships, task identity,
parameters, seed and runtime contract. Inspect
`qc/<comparison>_content_cache.tsv` for its recorded hit or miss reason. A
miss is not an analysis failure.

## Recover after a failure without replacing output

If a task fails, retain its configuration, logs and staging tree for diagnosis.
Do not rename the staging tree as a result and do not delete an earlier output
to make room for a retry. lisaR does not resume interrupted runs:
`pipeline.resume: true` stops with `LISA-RESUME-001`. A completed final output
directory is immutable and cannot be reused as a destination.

Prepare a reviewed retry beside the original configuration. Use one worker if
you encountered the contrast-evidence limitation above; otherwise keep a
worker request supported by your allocation. Always give the retry a new run
ID and output directory:

```{r parallel-retry-template, eval=FALSE}
retry <- yaml::read_yaml("study.yml")
retry$pipeline$workers <- 1L
retry$pipeline$run_id <- "study-retry-01"
retry$pipeline$output_dir <- "results/study-retry-01"
retry$pipeline$resume <- FALSE

retry_config <- "study-retry-01.yml"
stopifnot(!file.exists(retry_config))
yaml::write_yaml(retry, retry_config)

validation <- lisaR::validate_lisa_config(retry_config, check_files = TRUE)
stopifnot(validation$execution_ready)
plan <- lisaR::plan_lisa_outputs(retry_config)
plan$summary
```

Review the validation and plan before calling `run_lisa(retry_config)`. This
starts a new run; it does not continue or mutate the failed attempt.

```{r parallel-verify-template, eval=FALSE}
check <- lisaR::verify_lisa_run("results/study-retry-01")
stopifnot(identical(check$gate, "PASS"))
```

Verification applies to a completed run only. A dry-run planning directory is
not a scientific result and returns a failed verification gate by design.
