---
title: "Prepare DESeq2, limma or edgeR results for lisaR"
lang: en-GB
output:
  rmarkdown::html_vignette:
    mathjax: null
    includes:
      in_header: vignette-layout.inc
vignette: >
  %\VignetteIndexEntry{Prepare DESeq2, limma or edgeR results for lisaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Start with the complete result table from a differential-expression model that
you have already fitted. This guide shows two routes:

- use your exported table directly and name its columns in the lisaR study
  file; or
- use `prepare_lisa_de_input()` to create standard column names and a
  row-level audit.

The adapter is optional. It does not fit a model, choose a comparison or prove
which software produced a table.

The worked tables below are small fictional examples. The gene names make the
code easy to follow, but the numbers are not measurements. Use your complete
tested table for a real study.

## Choose the direct or adapted route

Your input needs one row per gene and these five values:

| Value | What to supply |
|---|---|
| Gene identifier | A non-empty identifier that matches the TERM2GENE membership resource. |
| Effect | The model's effect for the chosen comparison, such as `log2FoldChange` or `logFC`. |
| Signed ranking statistic | A statistic for the same comparison, ordered towards the declared positive direction. |
| P-value | The unadjusted p-value from the fitted model. |
| Adjusted p-value | The model's adjusted p-value for the complete tested population. |

If the exported table already contains these values, save it as TSV and point
`symbol_col`, `logfc_col`, `rank_col`, `pvalue_col` and `padj_col` to
the existing names in the study file. A DESeq2 Wald `stat` or a limma
moderated `t` can be used directly. You do not need to create
`rank_value`.

Use the adapter when standard names or its audit will make the hand-off
clearer. It returns `symbol`, `log2FoldChange`, `rank_value`, `pvalue`
and `padj`, ready to be named in a study file. The adapter works on an
exported data frame without loading DESeq2, limma or edgeR.

## State the comparison and positive direction

Before saving either form, write down the model comparison and what a positive
effect means. The signed statistic must describe the same comparison. Keep the
producing software, model specification and original export with the prepared
file.

| Input | Suitable starting point | What not to infer |
|---|---|---|
| DESeq2 Wald result | Signed `stat` for the stated coefficient/contrast | A DESeq2 likelihood-ratio `stat` need not be signed |
| limma one-contrast result | Signed moderated `t` | A multi-coefficient F statistic does not have one direction |
| edgeR one-contrast result | An explicitly declared signed ranking derived for that contrast | The adapter cannot infer direction from unsigned F/LR |
| Any result | Complete tested table, with original adjusted p-values | A significant-only table is not the GSEA universe |

Use the complete tested table rather than filtering to `padj < 0.05` before
GSEA. Non-significant genes remain part of the ordered background against
which a set is evaluated. Retain the adjusted p-values calculated for the
original tested population.

## Adapt a DESeq2-style Wald table

This fictional result has six tested genes. One has no statistic and another
has a finite ranking but no adjusted p-value, illustrating why absence of an
adjusted p-value should not silently remove a row from the saved table.

```{r deseq2-table, eval=TRUE}
library(lisaR)

wald <- data.frame(
  gene = c("CXCL9", "GZMB", "MKI67", "TCF7", "IL7R", "CCR7"),
  log2FoldChange = c(1.4, 0.8, -0.7, 0.15, -0.2, NA_real_),
  stat = c(4.2, 2.1, -2.8, 0.4, -0.6, NA_real_),
  pvalue = c(0.00003, 0.036, 0.005, 0.69, 0.55, NA_real_)
)
wald$padj <- p.adjust(wald$pvalue, method = "BH")
wald$padj[5L] <- NA_real_  # Illustrative missing adjusted p-value.

prepared <- prepare_lisa_de_input(
  wald,
  method = "DESeq2",
  columns = list(symbol = "gene"),
  positive_direction = "treated relative to control",
  matrix_scale = "transformed"
)
prepared$data

stopifnot(
  nrow(prepared$data) == nrow(wald),
  identical(prepared$data$rank_value, wald$stat),
  is.na(prepared$data$padj[5L]),
  prepared$row_audit$rank_eligible[5L],
  !prepared$row_audit$rank_eligible[6L]
)
```

The output columns are `symbol`, `log2FoldChange`, `rank_value`, `pvalue` and
`padj`. No threshold has been applied. The missing statistic remains missing;
it is not replaced by zero. `rank_eligible` records a finite ranking value,
not biological validity or every downstream eligibility condition.

Inspect what the adapter actually declared:

```{r input-audit, eval=TRUE}
prepared$column_mapping
prepared$row_audit[c(
  "source_row", "source_id", "retained", "rank_eligible",
  "missing_rank", "missing_padj"
)]
prepared$receipt[c(
  "declared_method", "positive_direction", "rank_source",
  "significance_filter", "producing_software_verified"
)]
```

`producing_software_verified` is false because a data frame cannot establish
which software fitted its model. Retain the original table and model records;
the adapter's receipt is an input declaration, not independent provenance.

For real DESeq2 results, export the complete result for the intended contrast,
then convert it as above. Do not combine a statistic for one coefficient with
an effect from another. For likelihood-ratio results, choose and justify a
compatible signed statistic separately rather than relabelling an unsigned
statistic.

## Adapt a limma result with its moderated t statistic

For one specified limma coefficient or contrast, the conventional effect and
p-value columns are recognised. The gene column is still declared explicitly:

```{r limma-table, eval=TRUE}
limma_table <- data.frame(
  gene = c("CXCL9", "GZMB", "MKI67", "TCF7"),
  logFC = c(1.3, 0.7, -0.6, 0.1),
  t = c(4.8, 2.5, -3.2, 0.3),
  P.Value = c(0.0001, 0.018, 0.004, 0.77)
)
limma_table$adj.P.Val <- p.adjust(limma_table$P.Value, "BH")
limma_input <- prepare_lisa_de_input(
  limma_table, method = "limma",
  columns = list(symbol = "gene"),
  positive_direction = "treated relative to control"
)
stopifnot(
  identical(limma_input$data$rank_value, limma_table$t),
  nrow(limma_input$data) == nrow(limma_table)
)
limma_input$data
```

When exporting a real `topTable()`, request all tested rows rather than its
default small top table, and use the intended single contrast. If biological
identifiers are in non-default row names, declare
`columns = list(symbol = ".rownames")`. Default row numbers are not gene IDs.

## Declare a signed edgeR ranking

An edgeR F or likelihood-ratio statistic is unsigned. For a **single
one-degree-of-freedom contrast**, one possible ranking convention is
`sign(logFC) * sqrt(F)`, using the effect for that same contrast. This is a
declared ranking transformation, not a new p-value or a claim that it is a
limma moderated t statistic. The design and test metadata must establish the
one-contrast restriction; the numbers in an exported table cannot prove it.

The following fictional example explicitly assumes that restriction:

```{r edger-table, eval=TRUE}
edger_table <- data.frame(
  symbol = c("CXCL9", "GZMB", "MKI67", "TCF7"),
  logFC = c(1.2, 0.6, -0.5, 0.1),
  F = c(16, 6.25, 9, 0.09),
  PValue = c(0.0008, 0.02, 0.006, 0.77)
)
edger_table$FDR <- p.adjust(edger_table$PValue, "BH")
edger_table$signed_sqrt_F <-
  sign(edger_table$logFC) * sqrt(edger_table$F)

edger_input <- prepare_lisa_de_input(
  edger_table, method = "edgeR",
  columns = list(rank = "signed_sqrt_F"),
  positive_direction = "treated relative to control"
)
stopifnot(
  identical(edger_input$data$rank_value, edger_table$signed_sqrt_F),
  identical(edger_input$data$padj, edger_table$FDR)
)
edger_input$data
```

For a multi-degree-of-freedom F or LR test, **do not attach the sign of an
arbitrarily chosen logFC**. Return to the fitted model and define a meaningful
one-dimensional contrast, or choose a method appropriate to the original
multidimensional question. Renaming `F` to `signed_rank` does not make it signed.

The adapter requires the rank mapping for edgeR and rejects conventional
unsigned F/LR columns and p-value/FDR columns as rankings. It also checks sign
agreement where both the rank and effect are finite and nonzero. Those checks
detect some mistakes; they cannot establish that a declared transformation was
scientifically justified.

## Resolve orientation, missing values and duplicates

### Orientation

`positive_direction` describes the effect in words. With the default
`rank_direction = "same_as_effect"`, the ranking must follow it too. If the
ranking really encodes the reverse comparison while the effect is correct,
`rank_direction = "opposite_to_effect"` negates the rank once and records the
choice. It does not reverse the effect, rename a contrast or repair mixed
model coefficients.

### Missing values

The default `na_policy = "retain"` saves missing statistics and records their
status. `na_policy = "error"` instead rejects them. Missing/empty identifiers,
infinite values and malformed numbers always fail. Keep model-specific reasons
such as filtering, non-convergence or independent filtering beside the source
table; the adapter does not reconstruct them.

### Duplicate identifiers

The default is an error, including case-insensitive collisions. Investigate
whether two rows are transcripts, aliases, repeated exports or distinct
features mapped to the same gene. Do not average their logFC, statistics or
p-values.

An explicit row-selection policy is available when it is justified. This toy
example demonstrates its audit, not a recommended biological rule:

```{r duplicate-audit, eval=TRUE}
duplicated_source <- rbind(wald, wald[1L, , drop = FALSE])
selected <- prepare_lisa_de_input(
  duplicated_source, "DESeq2",
  columns = list(symbol = "gene"),
  positive_direction = "treated relative to control",
  duplicate_policy = list(
    type = "select", criterion = "max_abs_rank", tie_breaker = "row_number"
  )
)
stopifnot(
  nrow(selected$full_data) == 7L,
  nrow(selected$data) == 6L,
  sum(!selected$row_audit$retained) == 1L
)
selected$row_audit[c("source_row", "source_id", "retained")]
```

The repeated row is preserved in `full_data` and its disposition is recorded.
On real non-identical duplicates, selecting the largest absolute statistic
can favour stronger effects. Choose the policy before interpretation and
retain the mapping/selection evidence. See the input contract in
[Configuration reference](configuration-reference.html) for row-level mapping
policies.

## Expression matrices are a separate input

The adapter only records `matrix_scale`; it neither reads nor transforms a
matrix, and that argument is not a YAML configuration field. Use
`matrix_scale = "transformed"` in the adapter receipt for values already
transformed, such as VST, rlog or documented log2 expression. Record
`"count_like"` only for count-like values. This declaration does not itself
configure downstream heatmaps. Their matrix-path convention is separate; see
the configuration reference before attaching a matrix. Do not describe VST
values as raw counts or apply a second log transform by accident.

Match matrix feature IDs to DE IDs, review sample order/annotation, and keep the
transformation and filtering provenance. An effect/ranking table cannot
reconstruct sample-level expression. The standard gene-evidence view can show
DE values without an expression matrix; expression heatmaps need that separate
input.

## Save the adapted table and its audit

Write the prepared table and its audit into a new directory:

```{r save-prepared, eval=FALSE}
input_dir <- "prepared-de-inputs"
stopifnot(!file.exists(input_dir))
dir.create(input_dir)

for (part in c(
  "data", "full_data", "receipt", "column_mapping",
  "row_audit", "duplicate_resolution"
)) {
  write.table(
    prepared[[part]], file.path(input_dir, paste0(part, ".tsv")),
    sep = "\t", quote = FALSE, row.names = FALSE, na = "NA"
  )
}
```

Retain the original source table and model record too: extra columns such as
model diagnostics are not copied into the canonical data.

The new folder contains the canonical TSV used by lisaR and the audit files
that explain how it was made. Keep the original export and model record beside
this folder.

Point the analysis configuration at `data.tsv` with this column mapping. This
is a fragment to place in the complete study file created in the next guide:

```yaml
single_de:
  - analysis_id: treated_vs_control
    label: Treated versus control
    comparison: Treated minus control
    positive_direction: Higher in treated than control
    de_path: prepared-de-inputs/data.tsv
    species: Homo sapiens
    symbol_col: symbol
    logfc_col: log2FoldChange
    rank_col: rank_value
    pvalue_col: pvalue
    padj_col: padj
```

For the adapted route, the rank column is now `rank_value`, not the original
`stat` or `t`. For the direct route, keep the original column name in
`rank_col`. Relative paths are interpreted from the folder containing the
configuration file.

The six-row example is too small for a real enrichment analysis. Apply the
chosen route to your complete tested results, then follow
[Getting started](getting-started.html) to validate resources, plan,
run and verify the study.

For supported DESeq2/edgeR objects already in memory, the alternative local API
is documented in [Function reference](function-reference.html). It is
separate from this table-first workflow; no object file needs to be
deserialised to follow this page.
