---
title: "Semantic categories and supercategories"
lang: en-GB
output:
  rmarkdown::html_vignette:
    mathjax: null
    includes:
      in_header: vignette-layout.inc
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Semantic categories and supercategories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use this guide to find one of the 158 LISA categories and understand the
biological scope in which the fixed dictionary places it. The catalogue
records dictionary assignments and representative member gene sets, not the
enrichment result of a particular study.

## Search for a category

In R or RStudio, read the installed catalogue as an ordinary data frame:

```{r semantic-lookup, eval=FALSE}
catalog_path <- system.file(
  "extdata", "semantic-catalog", "semantic_category_catalog.tsv",
  package = "lisaR", mustWork = TRUE
)
semantic_catalog <- read.delim(
  catalog_path, sep = "\t", quote = "", comment.char = "",
  colClasses = "character", check.names = FALSE
)

View(semantic_catalog)
```

Filter by words in the display name or description when you do not yet know an
ID:

```{r semantic-search, eval=FALSE}
query <- "repair|checkpoint"
matches <- grepl(
  query,
  paste(
    semantic_catalog$category_display_name,
    semantic_catalog$category_description
  ),
  ignore.case = TRUE
)
semantic_catalog[matches, c(
  "universe", "category_id", "category_display_name",
  "macrogroup_name", "core_member_gene_sets"
)]
```

Once you have an exact ID, select it together with its collection. This avoids
matching a similar label in the wrong biological scope:

```{r semantic-select, eval=FALSE}
chosen <- semantic_catalog[
  semantic_catalog$universe == "GOBP-C2" &
    semantic_catalog$category_id == "DNA_DAMAGE_REPAIR_CHECKPOINTS",
  c("category_display_name", "macrogroup_name", "core_member_gene_sets",
    "example_gene_set_1", "example_gene_set_2", "example_gene_set_3")
]
stopifnot(nrow(chosen) == 1L)
print(chosen, row.names = FALSE)
```

`View()` opens the data frame in RStudio's viewer; the other examples print
subsets in the R console. The count is membership in the core dictionary. The
one to three examples illustrate the category's scope. They are not its full
membership list or the strongest results from your study. The report's
evidence sheet, not this catalogue, supplies NES, FDR, leading edges and
gene-level DE.

## Member sets and category assignments

A gene set may be assigned to two categories in one supercategory. It then
contributes **one distinct set but two assignments**. The bundled count table
lets you check that distinction directly:

```{r semantic-distinct-support, eval=FALSE}
semantic_counts <- read.delim(system.file(
  "extdata", "semantic-catalog", "semantic_supercategory_counts.tsv",
  package = "lisaR", mustWork = TRUE
), sep = "\t", check.names = FALSE)
pi3k <- semantic_counts[
  semantic_counts$macrogroup_id == "PI3K_AKT_MTOR",
  c("universe", "macrogroup_id", "core_distinct_gene_sets",
    "core_member_assignments")
]
stopifnot(nrow(pi3k) == 1L,
          pi3k$core_distinct_gene_sets == 59L,
          pi3k$core_member_assignments == 88L)
print(pi3k, row.names = FALSE)
```

The 59 distinct gene sets have 88 assignments across categories in this
supercategory. Assignment counts and distinct gene-set counts therefore
measure different properties of the dictionary.

To open the corresponding study evidence, use the category navigator or
gene search described in the [report guide](scientific-report.html).

## Catalogue sources and scope

The four semantic collections are `GOBP-C2`, `GOMF`, `GOCC` and `PATHWAYS`.
`HALLMARKS` is a separate direct gene-set collection, not another LISA category
catalogue. Category descriptions and one to three exact member identifiers
were checked against `lisa_dictionary_core@1.0.0`. The existing provenance also
checks the same category vocabulary in
`lisa_dictionary_expanded@1.0.0`; this guide does not compare tiers or study
results.

The biological summary before each table describes the supercategory. Within a
table, read the exact category label and representative member IDs together.
The installed catalogue also includes a `category_description` field for text
searches. Its descriptions restate the category labels.

Descriptions and examples delimit a theme; they do not assert activation,
inhibition or causality. Example selection uses semantic representativeness
and stable tie-breaking on specificity, lexical match and gene-set identifier.
Dictionary membership supplies no numerical weight or gene-priority score.
Study evidence uses the member gene sets' enrichment and gene-level results.

Unassigned gene sets remain in the complete tested universe. The
`OTHER_UNCLASSIFIED` residual is tabular, not a biological category or figure.
For `GOBP-C2`, `GOMF` and `GOCC` it has dedicated residual tables; `PATHWAYS`
retains the complete GSEA ledger without a separate residual table.

The core and expanded classification dictionaries and category map are
included in lisaR. Gene-set-to-gene memberships are a separate external
resource. Source conditions are recorded in `THIRD_PARTY_NOTICES.md`. See
[Resources and provenance](resources-and-provenance.html) for those resource
roles; no additional resource installation is needed to read this catalogue.

Source versions and checksums accompany the installed catalogue in
`extdata/semantic-catalog/SEMANTIC_CATALOG_PROVENANCE.tsv`.

The active category map contains only categories with at least one exact
assignment in the core or expanded dictionary. Supercategories without a
retained category are absent from the active map.

| Collection | Supercategories | Categories | Core member assignments |
|---|---:|---:|---:|
| `GOBP-C2` | 10 | 40 | 11729 |
| `GOMF` | 9 | 37 | 1860 |
| `GOCC` | 6 | 12 | 1003 |
| `PATHWAYS` | 17 | 69 | 921 |

# Catalogue

## `GOBP-C2`

Use this collection for biological processes and programmes. It combines Gene
Ontology Biological Process with selected curated C2 pathway families, so one
supercategory can span cellular mechanisms, tissues and organismal
development. Read the exact member IDs before narrowing a broad category to
one experimental mechanism.

### Developmental and neural programs (`DEVELOPMENT_NEURAL`)

Groups developmental, morphogenetic, neurogenic and synaptic biological programmes.

Verified membership: 2 categories, 1738 distinct core gene sets and 1801 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Development and morphogenesis (`DEVELOPMENT_MORPHOGENESIS`) | 1010 | `GOBP_DEVELOPMENTAL_GROWTH_INVOLVED_IN_MORPHOGENESIS`<br>`GOBP_EAR_DEVELOPMENT`<br>`WP_HEART_DEVELOPMENT` |
| Neurogenesis and synapse (`NEUROGENESIS_SYNAPSE`) | 791 | `GOBP_NEUROGENESIS`<br>`GOBP_REGULATION_OF_NEUROGENESIS`<br>`GOBP_SYNAPSE_PRUNING` |

### Genome and gene expression (`GENOME_GENE_EXPRESSION`)

Groups genome maintenance and the transcriptional, post-transcriptional and translational control of gene expression.

Verified membership: 4 categories, 1076 distinct core gene sets and 1090 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| DNA damage, repair and checkpoints (`DNA_DAMAGE_REPAIR_CHECKPOINTS`) | 295 | `GOBP_REGULATION_OF_DNA_DAMAGE_CHECKPOINT`<br>`GOBP_DNA_REPAIR`<br>`REACTOME_DNA_REPAIR` |
| RNA processing and RNP biology (`RNA_PROCESSING_SPLICING_RNP`) | 263 | `GOBP_RNA_PROCESSING`<br>`GOBP_MITOCHONDRIAL_RNA_PROCESSING`<br>`GOBP_RNA_3_END_PROCESSING` |
| Transcription, chromatin and epigenetics (`TRANSCRIPTION_CHROMATIN_EPIGENETICS`) | 416 | `GOBP_CHROMATIN_LOOPING`<br>`GOBP_MRNA_TRANSCRIPTION`<br>`GOBP_RRNA_TRANSCRIPTION` |
| Translation and ribosome (`TRANSLATION_RIBOSOME`) | 116 | `KEGG_RIBOSOME`<br>`GOBP_TRANSLATION`<br>`REACTOME_TRANSLATION` |

### Immune and inflammatory biology (`IMMUNE_INFLAMMATION`)

Groups adaptive, innate, cytokine, complement and antigen-presentation programmes in immune and inflammatory biology.

Verified membership: 4 categories, 1321 distinct core gene sets and 1356 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Adaptive immunity and antigen presentation (`IMMUNE_ADAPTIVE_ANTIGEN_MHC`) | 328 | `GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION`<br>`KEGG_ANTIGEN_PROCESSING_AND_PRESENTATION`<br>`GOBP_REGULATION_OF_ANTIGEN_PROCESSING_AND_PRESENTATION` |
| Cytokine and inflammatory signaling (`IMMUNE_CYTOKINE_INFLAMMATORY`) | 535 | `WP_CYTOKINES_AND_INFLAMMATORY_RESPONSE`<br>`BIOCARTA_INFLAM_PATHWAY`<br>`BIOCARTA_CYTOKINE_PATHWAY` |
| Immune core (`IMMUNE_CORE`) | 185 | `GOBP_REGULATION_OF_IMMUNE_RESPONSE`<br>`GOBP_IMMUNE_EFFECTOR_PROCESS`<br>`GOBP_ACTIVATION_OF_IMMUNE_RESPONSE` |
| Innate immunity, complement and PRR (`IMMUNE_INNATE_COMPLEMENT_PRR`) | 308 | `WP_COMPLEMENT_SYSTEM`<br>`WP_COMPLEMENT_ACTIVATION`<br>`GOBP_COMPLEMENT_ACTIVATION` |

### Metabolism and bioenergetics (`METABOLISM_BIOENERGETICS`)

Groups amino-acid, carbohydrate, lipid, nucleotide, mitochondrial and detoxification programmes that support metabolism and energy balance.

Verified membership: 8 categories, 1560 distinct core gene sets and 1576 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Amino acid and one-carbon metabolism (`METABOLISM_AA_ONECARBON`) | 251 | `WP_AMINO_ACID_METABOLISM`<br>`GOBP_AMINO_ACID_METABOLIC_PROCESS`<br>`GOBP_ONE_CARBON_METABOLIC_PROCESS` |
| Carbohydrate metabolism and glycolysis (`METABOLISM_CARBOHYDRATE_GLYCOLYSIS`) | 172 | `GOBP_CARBOHYDRATE_METABOLIC_PROCESS`<br>`GOBP_REGULATION_OF_CARBOHYDRATE_METABOLIC_PROCESS`<br>`GOBP_CARBOHYDRATE_DERIVATIVE_METABOLIC_PROCESS` |
| Detoxification and xenobiotic metabolism (`DETOX_XENOBIOTIC_METABOLISM`) | 126 | `GOBP_XENOBIOTIC_METABOLIC_PROCESS`<br>`GOBP_DETOXIFICATION`<br>`REACTOME_XENOBIOTICS` |
| General metabolism (`METABOLISM_CORE`) | 51 | `GOBP_ATP_METABOLIC_PROCESS`<br>`GOBP_RECEPTOR_METABOLIC_PROCESS`<br>`REACTOME_METABOLISM_OF_COFACTORS` |
| Lipid and fatty acid metabolism (`METABOLISM_LIPID_FA`) | 637 | `KEGG_FATTY_ACID_METABOLISM`<br>`REACTOME_FATTY_ACID_METABOLISM`<br>`GOBP_FATTY_ACID_METABOLIC_PROCESS` |
| Mitochondria, OXPHOS and ETC (`MITOCHONDRIA_OXPHOS_ETC`) | 143 | `BIOCARTA_ETC_PATHWAY`<br>`GOBP_MITOCHONDRIAL_FUSION`<br>`GOBP_MITOCHONDRIAL_FISSION` |
| Nucleotide metabolism (`METABOLISM_NUCLEOTIDE`) | 175 | `WP_NUCLEOTIDE_METABOLISM`<br>`GOBP_NUCLEOBASE_METABOLIC_PROCESS`<br>`GOBP_NUCLEOSIDE_METABOLIC_PROCESS` |
| Peroxisome and fatty acid oxidation (`PEROXISOME_FAO`) | 21 | `GOBP_FATTY_ACID_ALPHA_OXIDATION`<br>`KEGG_PEROXISOME`<br>`GOBP_FATTY_ACID_BETA_OXIDATION_USING_ACYL_COA_OXIDASE` |

### Microenvironment, vascular and ECM (`MICROENVIRONMENT_VASCULAR_ECM`)

Groups extracellular-matrix remodelling, vascular, angiogenic and lymphatic programmes in the tissue microenvironment.

Verified membership: 2 categories, 333 distinct core gene sets and 334 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| ECM and matrix remodeling (`ECM_MATRIX_REMODELING`) | 183 | `REACTOME_ECM_PROTEOGLYCANS`<br>`WP_MATRIX_METALLOPROTEINASES`<br>`KEGG_ECM_RECEPTOR_INTERACTION` |
| Vascular, angiogenic and lymphatic biology (`VASCULAR_LYMPHATIC`) | 151 | `WP_ANGIOGENESIS`<br>`GOBP_VASCULOGENESIS`<br>`GOBP_REGULATION_OF_VASCULOGENESIS` |

### Proliferation and cell cycle (`PROLIFERATION_CELL_CYCLE`)

Groups programmes governing DNA replication, mitosis and cell proliferation.

Verified membership: 1 category, 555 distinct core gene sets and 555 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Cell cycle and proliferation (`CELL_CYCLE_PROLIFERATION`) | 555 | `WP_CELL_CYCLE`<br>`GOBP_CELL_CYCLE`<br>`KEGG_CELL_CYCLE` |

### Signaling and cell communication (`SIGNALING_COMMUNICATION`)

Groups broad growth-factor, hormonal and developmental pathways used for intercellular communication.

Verified membership: 5 categories, 1964 distinct core gene sets and 2015 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Core cell signaling (`SIGNALING_CORE`) | 714 | `GOBP_CELL_CELL_SIGNALING`<br>`GOBP_CELL_ACTIVATION`<br>`GOBP_NEGATIVE_REGULATION_OF_CELL_COMMUNICATION` |
| Endocrine and hormonal pathways (`ENDOCRINE_HORMONAL`) | 497 | `GOBP_ENDOCRINE_HORMONE_SECRETION`<br>`GOBP_ENDOCRINE_PROCESS`<br>`GOBP_REGULATION_OF_ENDOCRINE_PROCESS` |
| Growth factor and RTK pathways (`GROWTH_FACTOR_RTK`) | 412 | `GOBP_RESPONSE_TO_GROWTH_FACTOR`<br>`GOBP_NERVE_GROWTH_FACTOR_SIGNALING_PATHWAY`<br>`GOBP_GROWTH` |
| TGF-beta and SMAD signaling (`TGFB_SMAD`) | 129 | `KEGG_TGF_BETA_SIGNALING_PATHWAY`<br>`REACTOME_TGFBR3_REGULATES_TGF_BETA_SIGNALING`<br>`KEGG_MEDICUS_REFERENCE_TGF_BETA_SIGNALING_PATHWAY` |
| Wnt, Notch, Hedgehog and Hippo (`WNT_NOTCH_HEDGEHOG_HIPPO`) | 263 | `WNT_SIGNALING`<br>`WP_WNT_SIGNALING`<br>`PID_NOTCH_PATHWAY` |

### Stress, proteostasis and cell death (`STRESS_DEATH_PROTEOSTASIS`)

Groups stress adaptation, autophagy, proteostasis and regulated cell-death programmes.

Verified membership: 7 categories, 879 distinct core gene sets and 901 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Apoptosis and regulated cell death (`APOPTOSIS_CELL_DEATH`) | 321 | `SA_PROGRAMMED_CELL_DEATH`<br>`GOBP_PYROPTOTIC_CELL_DEATH`<br>`GOBP_B_CELL_APOPTOTIC_PROCESS` |
| Autophagy, lysosome and phagosome maturation (`AUTOPHAGY_LYSOSOME`) | 154 | `GOBP_PHAGOSOME_MATURATION`<br>`GOBP_AUTOPHAGOSOME_MATURATION`<br>`GOBP_REGULATION_OF_AUTOPHAGOSOME_MATURATION` |
| ER stress and unfolded protein response (`ER_STRESS_UPR`) | 48 | `WP_UNFOLDED_PROTEIN_RESPONSE`<br>`REACTOME_UNFOLDED_PROTEIN_RESPONSE_UPR`<br>`GOBP_CELLULAR_RESPONSE_TO_UNFOLDED_PROTEIN` |
| Heat shock and acute stress response (`HEAT_SHOCK_STRESS_RESPONSE`) | 17 | `REACTOME_CELLULAR_RESPONSE_TO_HEAT_STRESS`<br>`REACTOME_REGULATION_OF_HSF1_MEDIATED_HEAT_SHOCK_RESPONSE` |
| Hypoxia and HIF programs (`HYPOXIA_HIF`) | 31 | `BIOCARTA_HIF_PATHWAY`<br>`REACTOME_CELLULAR_RESPONSE_TO_HYPOXIA`<br>`GOBP_REGULATION_OF_CELLULAR_RESPONSE_TO_HYPOXIA` |
| Oxidative stress, redox and ROS (`OXIDATIVE_STRESS_REDOX_ROS`) | 116 | `WP_OXIDATIVE_STRESS_RESPONSE`<br>`GOBP_RESPONSE_TO_OXIDATIVE_STRESS`<br>`GOBP_REGULATION_OF_RESPONSE_TO_OXIDATIVE_STRESS` |
| Proteostasis, ubiquitin and proteasome (`PROTEOSTASIS_UBIQUITIN_PROTEASOME`) | 214 | `KEGG_UBIQUITIN_MEDIATED_PROTEOLYSIS`<br>`KEGG_PROTEASOME`<br>`GOBP_PROTEOLYSIS` |

### Structure, motility and plasticity (`STRUCTURE_MOTILITY_PLASTICITY`)

Groups adhesion, polarity, cytoskeletal, migratory and cell-state-plasticity programmes.

Verified membership: 4 categories, 911 distinct core gene sets and 928 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Adhesion, junctions and polarity (`ADHESION_JUNCTION_POLARITY`) | 260 | `KEGG_GAP_JUNCTION`<br>`WP_FOCAL_ADHESION`<br>`GOBP_CELL_ADHESION` |
| Cytoskeleton and contractility (`CYTOSKELETON_CONTRACTILITY`) | 419 | `GOBP_MUSCLE_CONTRACTION`<br>`REACTOME_MUSCLE_CONTRACTION`<br>`GOBP_CYTOSKELETON_ORGANIZATION` |
| EMT and cell-state plasticity (`EMT_PLASTICITY`) | 34 | `GOBP_MESENCHYMAL_CELL_MIGRATION`<br>`GOBP_MESENCHYMAL_CELL_DIFFERENTIATION`<br>`GOBP_NEURAL_CREST_CELL_DIFFERENTIATION` |
| Migration, invasion and motility programs (`MIGRATION_INVASION`) | 215 | `GOBP_CELL_MOTILITY`<br>`GOBP_CELL_MIGRATION`<br>`GOBP_NEURON_MIGRATION` |

### Transport, trafficking and homeostasis (`TRANSPORT_TRAFFICKING_HOMEOSTASIS`)

Groups membrane transport, localisation, vesicle traffic and ionic homeostasis programmes.

Verified membership: 3 categories, 1150 distinct core gene sets and 1173 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Ion transport and homeostasis (`ION_TRANSPORT_HOMEOSTASIS`) | 382 | `REACTOME_ION_HOMEOSTASIS`<br>`GOBP_IRON_ION_TRANSPORT`<br>`GOBP_ZINC_ION_TRANSPORT` |
| Transport and localization (`TRANSPORT_LOCALIZATION`) | 418 | `GOBP_REGULATION_OF_TRANSPORT`<br>`GOBP_ADP_TRANSPORT`<br>`GOBP_ATP_TRANSPORT` |
| Vesicle, endocytosis, exocytosis and secretion (`VESICLE_SECRETORY_TRAFFICKING`) | 373 | `GOBP_SYNAPTIC_VESICLE_EXOCYTOSIS`<br>`GOBP_ACROSOMAL_VESICLE_EXOCYTOSIS`<br>`GOBP_REGULATION_OF_SYNAPTIC_VESICLE_EXOCYTOSIS` |

## `GOMF`

Use this collection for the molecular activity performed by a gene product,
such as binding, catalysis, transport or receptor activity. It does not by
itself say which broader biological process is active. Categories can overlap
when one GO molecular-function set fits more than one functional grouping.

### Catalytic and metabolic enzyme functions (`MF_CATALYTIC_METABOLIC`)

Groups molecular functions carried out by metabolic, redox, detoxification and other catalytic enzymes.

Verified membership: 7 categories, 358 distinct core gene sets and 368 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Amino-acid metabolic enzyme activity (`MF_AMINO_ACID_METAB`) | 7 | `GOMF_ACID_AMINO_ACID_LIGASE_ACTIVITY`<br>`GOMF_L_AMINO_ACID_N_ACETYLTRANSFERASE_ACTIVITY` |
| Antioxidant and redox activity (`MF_ANTIOX_REDOX`) | 54 | `GOMF_ANTIOXIDANT_ACTIVITY`<br>`GOMF_PEROXIREDOXIN_ACTIVITY`<br>`GOMF_THIOL_OXIDASE_ACTIVITY` |
| Carbohydrate metabolic enzyme activity (`MF_CARBOHYDRATE_METAB`) | 58 | `GOMF_CARBOHYDRATE_KINASE_ACTIVITY`<br>`GOMF_CARBOHYDRATE_PHOSPHATASE_ACTIVITY`<br>`GOMF_RACEMASE_AND_EPIMERASE_ACTIVITY_ACTING_ON_CARBOHYDRATES_AND_DERIVATIVES` |
| Detoxification and xenobiotic activity (`MF_DETOX_XENOBIOTIC`) | 6 | `GOMF_ARYL_SULFOTRANSFERASE_ACTIVITY`<br>`GOMF_ESTROGEN_2_HYDROXYLASE_ACTIVITY` |
| General catalytic activity (`MF_CATALYTIC`) | 134 | `GOMF_CATALYTIC_ACTIVITY_ACTING_ON_DNA`<br>`GOMF_CATALYTIC_ACTIVITY_ACTING_ON_RNA`<br>`GOMF_CATALYTIC_ACTIVITY_ACTING_ON_A_RRNA` |
| General metabolic enzyme activity (`MF_METABOLIC_ENZ`) | 23 | `GOMF_COA_TRANSFERASE_ACTIVITY`<br>`GOMF_SULFOTRANSFERASE_ACTIVITY`<br>`GOMF_SULFURTRANSFERASE_ACTIVITY` |
| Lipid metabolic enzyme activity (`MF_LIPID_METAB`) | 86 | `GOMF_LIPASE_ACTIVITY`<br>`GOMF_DEACYLASE_ACTIVITY`<br>`GOMF_COA_LIGASE_ACTIVITY` |

### Cell-cycle and developmental regulatory functions (`MF_CELL_CYCLE_DEVELOPMENT`)

Groups molecular regulators of cell-cycle progression, differentiation and developmental decisions.

Verified membership: 2 categories, 4 distinct core gene sets and 4 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Cell-cycle regulatory activity (`MF_CELL_CYCLE_REG`) | 3 | `GOMF_CYCLIN_DEPENDENT_PROTEIN_KINASE_REGULATOR_ACTIVITY` |
| Developmental regulator activity (`MF_DEVELOPMENT_REG`) | 1 | `GOMF_MORPHOGEN_ACTIVITY` |

### Gene-expression molecular functions (`MF_GENE_EXPRESSION`)

Groups chromatin, transcription, RNA-processing and translation-related molecular functions.

Verified membership: 4 categories, 170 distinct core gene sets and 170 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Epigenetic and chromatin activity (`MF_EPIGENETIC`) | 73 | `GOMF_CHROMATIN_DNA_BINDING`<br>`GOMF_CHROMATIN_PROTEIN_ADAPTOR_ACTIVITY`<br>`GOMF_PROMOTER_SPECIFIC_CHROMATIN_BINDING` |
| RNA processing activity (`MF_RNA_PROCESSING`) | 47 | `GOMF_RNA_NUCLEASE_ACTIVITY`<br>`GOMF_RNA_POLYMERASE_ACTIVITY`<br>`GOMF_RNA_EXONUCLEASE_ACTIVITY` |
| Transcription regulation activity (`MF_TRANSCRIPTION_REG`) | 41 | `GOMF_TRANSCRIPTION_REGULATOR_ACTIVITY`<br>`GOMF_TRANSCRIPTION_COACTIVATOR_ACTIVITY`<br>`GOMF_TRANSCRIPTION_COREGULATOR_ACTIVITY` |
| Translation regulation activity (`MF_TRANSLATION_REG`) | 9 | `GOMF_TRANSLATION_FACTOR_ACTIVITY`<br>`GOMF_TRANSLATION_ACTIVATOR_ACTIVITY` |

### Generic or host-interaction functions (`MF_GENERIC_OTHER`)

Groups broad binding or host-interaction molecular functions that do not fit a more specific functional class.

Verified membership: 2 categories, 525 distinct core gene sets and 526 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Binding activity (`MF_BINDING`) | 525 | `GOMF_ADP_BINDING`<br>`GOMF_AMP_BINDING`<br>`GOMF_BMP_BINDING` |
| Viral-host molecular function (`MF_VIRAL_HOST`) | 1 | `GOMF_VIRION_BINDING` |

### Immune, ligand and receptor functions (`MF_IMMUNE_LIGAND_RECEPTOR`)

Groups immune recognition, complement, receptor and extracellular ligand molecular functions.

Verified membership: 3 categories, 53 distinct core gene sets and 53 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Complement molecular function (`MF_COMPLEMENT`) | 3 | `GOMF_COMPLEMENT_BINDING` |
| Cytokine, chemokine, growth-factor and hormone ligand activity (`MF_CYTOKINE_GROWTH`) | 12 | `GOMF_GROWTH_FACTOR_ACTIVITY`<br>`GOMF_HORMONE_ACTIVITY` |
| Immune receptor and antigen-recognition activity (`MF_IMMUNE_RECEPTOR`) | 38 | `GOMF_IMMUNE_RECEPTOR_ACTIVITY`<br>`GOMF_PATTERN_RECOGNITION_RECEPTOR_ACTIVITY`<br>`GOMF_PEPTIDOGLYCAN_IMMUNE_RECEPTOR_ACTIVITY` |

### Proteostasis, stress and cell-death functions (`MF_PROTEOSTASIS_STRESS_DEATH`)

Groups molecular functions involved in proteostasis, stress responses, proteolysis and regulated cell death.

Verified membership: 5 categories, 82 distinct core gene sets and 84 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Apoptosis and cell-death regulatory activity (`MF_APOPTOSIS_REG`) | 3 | `GOMF_DEATH_RECEPTOR_ACTIVITY` |
| Protease and protease inhibitor activity (`MF_PROTEASE_INHIB`) | 40 | `GOMF_PEPTIDASE_INHIBITOR_ACTIVITY`<br>`GOMF_METALLOENDOPEPTIDASE_INHIBITOR_ACTIVITY`<br>`GOMF_MOLECULAR_FUNCTION_INHIBITOR_ACTIVITY` |
| Protein folding and degradation activity (`MF_PROTEIN_FOLDING_DEG`) | 11 | `GOMF_PROTEIN_FOLDING_CHAPERONE`<br>`GOMF_PROTEIN_FOLDING_CHAPERONE_BINDING` |
| Stress-response molecular activity (`MF_STRESS_RESPONSE`) | 2 | `GOMF_HEAT_SHOCK_PROTEIN_BINDING` |
| Ubiquitin and SUMO ligase activity (`MF_LIGASE_SUMOYL`) | 28 | `GOMF_SUMO_LIGASE_ACTIVITY`<br>`GOMF_UBIQUITIN_UBIQUITIN_LIGASE_ACTIVITY`<br>`GOMF_HISTONE_UBIQUITIN_LIGASE_ACTIVITY` |

### Signaling and regulatory functions (`MF_SIGNALING_REGULATION`)

Groups receptors, kinases, phosphatases, adaptors and enzyme regulators that control signalling.

Verified membership: 5 categories, 306 distinct core gene sets and 313 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Enzyme regulator activity (`MF_ENZYME_REG`) | 59 | `GOMF_ENZYME_REGULATOR_ACTIVITY`<br>`GOMF_ATPASE_REGULATOR_ACTIVITY`<br>`GOMF_ENZYME_ACTIVATOR_ACTIVITY` |
| Intracellular signaling activity (`MF_SIGNALING`) | 12 | `GOMF_GTPASE_ACTIVITY`<br>`GOMF_G_PROTEIN_ACTIVITY` |
| Kinase and phosphatase activity (`MF_KINASE_PHOSPHATASE`) | 104 | `GOMF_PHOSPHATIDYLINOSITOL_KINASE_ACTIVITY`<br>`GOMF_MAP_KINASE_PHOSPHATASE_ACTIVITY`<br>`GOMF_INOSITOL_PHOSPHATE_KINASE_ACTIVITY` |
| Receptor activity (`MF_RECEPTOR`) | 122 | `GOMF_BMP_RECEPTOR_ACTIVITY`<br>`GOMF_WNT_RECEPTOR_ACTIVITY`<br>`GOMF_GABA_RECEPTOR_ACTIVITY` |
| Scaffold and adaptor activity (`MF_SCAFFOLD_ADAPTER`) | 16 | `GOMF_SIGNALING_ADAPTOR_ACTIVITY`<br>`GOMF_CARGO_ADAPTOR_ACTIVITY` |

### Structure, adhesion and ECM functions (`MF_STRUCTURE_ADHESION_ECM`)

Groups structural, motor, adhesion and extracellular-matrix molecular functions.

Verified membership: 4 categories, 54 distinct core gene sets and 55 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Adhesion molecular function (`MF_ADHESION`) | 12 | `GOMF_CELL_ADHESION_MOLECULE_BINDING`<br>`GOMF_CELL_ADHESION_MEDIATOR_ACTIVITY` |
| Cytoskeleton and motor activity (`MF_CYTOSKELETON_MOTOR`) | 13 | `GOMF_CYTOSKELETAL_MOTOR_ACTIVITY`<br>`GOMF_MICROTUBULE_MOTOR_ACTIVITY` |
| Extracellular matrix molecular function (`MF_EXTRACELL_MATRIX`) | 13 | `GOMF_EXTRACELLULAR_MATRIX_BINDING`<br>`GOMF_EXTRACELLULAR_MATRIX_PROTEIN_BINDING` |
| Structural constituent activity (`MF_STRUCTURAL_CONSTITUENT`) | 17 | `GOMF_STRUCTURAL_CONSTITUENT_OF_MUSCLE`<br>`GOMF_STRUCTURAL_CONSTITUENT_OF_SYNAPSE` |

### Transport, vesicle traffic and sensing functions (`MF_TRANSPORT_SENSING`)

Groups channels, pumps, transporters, vesicle-trafficking factors and physicochemical sensors.

Verified membership: 5 categories, 285 distinct core gene sets and 287 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Ion and physicochemical sensing activity (`MF_ION_SENSING`) | 4 | `GOMF_METAL_ION_SENSOR_ACTIVITY` |
| Ion channel and pump activity (`MF_ION_CHANNEL_PUMP`) | 86 | `GOMF_ATP_GATED_ION_CHANNEL_ACTIVITY`<br>`GOMF_POTASSIUM_ION_LEAK_CHANNEL_ACTIVITY`<br>`GOMF_MONOATOMIC_ION_GATED_CHANNEL_ACTIVITY` |
| Olfactory and taste receptor activity (`MF_OLFACTORY`) | 6 | `GOMF_TASTE_RECEPTOR_ACTIVITY`<br>`GOMF_OLFACTORY_RECEPTOR_ACTIVITY` |
| Transporter activity (`MF_TRANSPORTER`) | 188 | `GOMF_TRANSPORTER_ACTIVITY`<br>`GOMF_LIPID_TRANSPORTER_ACTIVITY`<br>`GOMF_PROTEIN_TRANSPORTER_ACTIVITY` |
| Vesicle trafficking activity (`MF_VESICLE_TRAFFIC`) | 3 | `GOMF_SNARE_BINDING` |

## `GOCC`

Use this collection for cellular locations, structures and complexes. A GOCC
category can support a compartment-level interpretation, but enrichment is not
direct evidence that a protein moved to that compartment or that the structure
changed in abundance.

### Cell surface, adhesion and ECM compartments (`CC_SURFACE_ADHESION_ECM`)

Covers plasma-membrane, cell-junction, cytoskeletal and extracellular-matrix compartments.

Verified membership: 3 categories, 343 distinct core gene sets and 348 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Extracellular matrix compartment (`CC_EXTRACELLULAR_MATRIX`) | 24 | `GOCC_SPECIALIZED_EXTRACELLULAR_MATRIX`<br>`GOCC_SYNAPSE_ASSOCIATED_EXTRACELLULAR_MATRIX`<br>`GOCC_INTERSTITIAL_MATRIX` |
| Junction, adhesion and cytoskeletal compartment (`CC_JUNCTION_ADHESION_CYTOSKELETON`) | 209 | `GOCC_GAP_JUNCTION`<br>`GOCC_TIGHT_JUNCTION`<br>`GOCC_ADHERENS_JUNCTION` |
| Plasma membrane and cell-surface compartment (`CC_PLASMA_MEMBRANE_CELL_SURFACE`) | 115 | `GOCC_CELL_SURFACE`<br>`GOCC_CELL_BODY_MEMBRANE`<br>`GOCC_PLASMA_MEMBRANE_RAFT` |

### Immune and signaling complexes (`CC_IMMUNE_COMPLEXES`)

Covers cellular complexes that assemble immune and inflammatory signalling machinery.

Verified membership: 1 category, 29 distinct core gene sets and 29 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Immune signaling and inflammatory complex compartment (`CC_IMMUNE_SIGNALING_COMPLEX`) | 29 | `GOCC_NLRP1_INFLAMMASOME_COMPLEX`<br>`GOCC_NLRP3_INFLAMMASOME_COMPLEX`<br>`GOCC_CANONICAL_INFLAMMASOME_COMPLEX` |

### Metabolic organelles (`CC_METABOLIC_ORGANELLES`)

Covers organelles that host energy production and intermediary metabolism, especially mitochondria and peroxisomes.

Verified membership: 1 category, 46 distinct core gene sets and 46 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Mitochondrial and metabolic organelle compartment (`CC_MITOCHONDRIAL_METABOLIC`) | 46 | `GOCC_MITOCHONDRION`<br>`GOCC_ORGANELLAR_RIBOSOME`<br>`GOCC_MITOCHONDRIAL_CRISTA` |

### Nuclear and gene-expression compartments (`CC_NUCLEAR_GENE_EXPRESSION`)

Covers nuclear, chromatin, transcriptional, RNA-processing and ribosomal compartments.

Verified membership: 2 categories, 275 distinct core gene sets and 277 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Nuclear and chromatin compartment (`CC_NUCLEAR_CHROMATIN`) | 137 | `GOCC_CHROMATIN`<br>`GOCC_NUCLEAR_BODY`<br>`GOCC_NUCLEAR_PORE` |
| Transcription, RNA and ribosome complexes (`CC_TRANSCRIPTION_RNA_RIBOSOME`) | 140 | `GOCC_RNA_DECAPPING_COMPLEX`<br>`GOCC_RNA_POLYMERASE_COMPLEX`<br>`GOCC_TRANSCRIPTION_EXPORT_COMPLEX` |

### Proteostasis and generic complexes (`CC_PROTEOSTASIS_GENERIC_COMPLEXES`)

Covers macromolecular complexes involved in protein folding, degradation and other general protein assemblies.

Verified membership: 2 categories, 76 distinct core gene sets and 76 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Other generic protein/macromolecular complex (`CC_OTHER_GENERIC_COMPLEX`) | 42 | `GOCC_PROTEIN_KINASE_COMPLEX`<br>`GOCC_MEMBRANE_PROTEIN_COMPLEX`<br>`GOCC_PROTEIN_PHOSPHATASE_4_COMPLEX` |
| Proteostasis complex compartment (`CC_PROTEOSTASIS_COMPLEX`) | 34 | `GOCC_3M_COMPLEX`<br>`GOCC_RQC_COMPLEX`<br>`GOCC_PREFOLDIN_COMPLEX` |

### Secretory and endomembrane compartments (`CC_SECRETORY_ENDOMEMBRANE`)

Covers endoplasmic-reticulum, Golgi, vesicular, endolysosomal and secretory compartments.

Verified membership: 3 categories, 226 distinct core gene sets and 227 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Endolysosome, autophagy and intracellular vesicle compartment (`CC_ENDOLYSOSOME_AUTOPHAGY_VESICLE`) | 116 | `GOCC_ENDOLYSOSOME`<br>`GOCC_AUTOPHAGOSOME`<br>`GOCC_VESICLE_COAT` |
| ER, Golgi and secretory-route compartment (`CC_ER_GOLGI_SECRETORY`) | 96 | `GOCC_ENDOPLASMIC_RETICULUM_GOLGI_INTERMEDIATE_COMPARTMENT`<br>`GOCC_GOLGI_LUMEN`<br>`GOCC_GOLGI_STACK` |
| Extracellular vesicle and secretome compartment (`CC_EXTRACELLULAR_VESICLE_SECRETOME`) | 15 | `GOCC_CHYLOMICRON`<br>`GOCC_MICROVESICLE` |

## `PATHWAYS`

Use this stricter signalling collection to distinguish named receptors,
branches and downstream cascades within the broader GOBP-C2 space. Categories
can share a gene set when a curated pathway crosses more than one branch. A
category label organises the saved enrichment evidence; it is not a claim that
the named pathway is activated.

### RTK/RAS/MAPK signaling (`RTK_RAS_MAPK`)

Groups receptor-tyrosine-kinase signalling through RAS and the RAF-MEK-ERK cascade.

Verified membership: 12 categories, 252 distinct core gene sets and 279 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| EGFR/ERBB signaling (`EGFR_ERBB_SIGNALING`) | 67 | `WP_ERBB_SIGNALING`<br>`GOBP_ERBB_SIGNALING_PATHWAY`<br>`KEGG_ERBB_SIGNALING_PATHWAY` |
| FGFR signaling (`FGFR_SIGNALING`) | 42 | `REACTOME_SIGNALING_BY_FGFR`<br>`KEGG_MEDICUS_REFERENCE_FGF_FGFR_PI3K_SIGNALING_PATHWAY`<br>`KEGG_MEDICUS_REFERENCE_FGF_FGFR_RAS_ERK_SIGNALING_PATHWAY` |
| VEGF/VEGFR signaling (`VEGF_VEGFR_SIGNALING`) | 11 | `PID_VEGF_VEGFR_PATHWAY`<br>`BIOCARTA_VEGF_PATHWAY` |
| PDGF/PDGFR signaling (`PDGF_PDGFRA_B_SIGNALING`) | 15 | `WP_PDGF_PATHWAY`<br>`BIOCARTA_PDGF_PATHWAY` |
| MET/HGF signaling (`MET_HGF_SIGNALING`) | 15 | `PID_MET_PATHWAY`<br>`BIOCARTA_MET_PATHWAY` |
| KIT signaling (`KIT_SIGNALING`) | 8 | `PID_KIT_PATHWAY`<br>`WP_KIT_RECEPTOR_SIGNALING` |
| RET signaling (`RET_SIGNALING`) | 6 | `PID_RET_PATHWAY`<br>`REACTOME_RET_SIGNALING` |
| Insulin/IGF signaling (`INSULIN_IGF_SIGNALING`) | 25 | `PID_INSULIN_PATHWAY`<br>`WP_INSULIN_SIGNALING`<br>`BIOCARTA_INSULIN_PATHWAY` |
| RAS activation (`RAS_ACTIVATION`) | 16 | `PID_RAS_PATHWAY`<br>`WP_RAS_SIGNALING` |
| RAF-MEK-ERK cascade (`RAF_MEK_ERK_CASCADE`) | 43 | `BIOCARTA_ERK_PATHWAY`<br>`WP_MAPK_CASCADE`<br>`REACTOME_RAF_ACTIVATION` |
| ALK signaling (`ALK_SIGNALING`) | 7 | `BIOCARTA_ALK_PATHWAY`<br>`REACTOME_SIGNALING_BY_ALK` |
| NTRK/TRK signaling (`NTRK_TRK_SIGNALING`) | 24 | `PID_MAPK_TRK_PATHWAY`<br>`PID_PI3K_PLC_TRK_PATHWAY`<br>`GOBP_NEUROTROPHIN_TRK_RECEPTOR_SIGNALING_PATHWAY` |

### Stress MAPK signaling (`STRESS_MAPK`)

Groups stress-responsive p38 and JNK mitogen-activated protein kinase pathways.

Verified membership: 2 categories, 21 distinct core gene sets and 21 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| p38 MAPK signaling (`P38_MAPK_SIGNALING`) | 12 | `WP_P38_MAPK_SIGNALING`<br>`PID_P38_MK2_PATHWAY` |
| JNK signaling (`JNK_SIGNALING`) | 9 | `GOBP_JNK_CASCADE`<br>`KEGG_MEDICUS_REFERENCE_FAS_JNK_SIGNALING_PATHWAY` |

### PI3K/AKT/mTOR signaling (`PI3K_AKT_MTOR`)

Groups PI3K, AKT, mTOR and AMPK pathways that coordinate growth, survival and nutrient sensing.

Verified membership: 5 categories, 59 distinct core gene sets and 88 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Class I PI3K signaling (`PI3K_CLASS_I_SIGNALING`) | 33 | `REACTOME_PI3K_AKT_ACTIVATION`<br>`KEGG_MEDICUS_REFERENCE_P4_PR_PI3K_SIGNALING_PATHWAY`<br>`KEGG_MEDICUS_REFERENCE_HGF_MET_PI3K_SIGNALING_PATHWAY` |
| AKT signaling (`AKT_SIGNALING`) | 37 | `BIOCARTA_AKT_PATHWAY`<br>`PID_PI3KCI_AKT_PATHWAY`<br>`REACTOME_PI3K_AKT_ACTIVATION` |
| mTORC1 signaling (`MTORC1_SIGNALING`) | 11 | `REACTOME_MTORC1_MEDIATED_SIGNALLING`<br>`REACTOME_AMINO_ACIDS_REGULATE_MTORC1` |
| mTORC2 signaling (`MTORC2_SIGNALING`) | 1 | `GOBP_TORC2_SIGNALING` |
| AMPK signaling (`AMPK_SIGNALING`) | 6 | `GOBP_CAMKK_AMPK_SIGNALING_CASCADE`<br>`REACTOME_ACTIVATION_OF_AMPK_DOWNSTREAM_OF_NMDARS` |

### RHO GTPase and cytoskeletal signaling (`RHO_GTPASE_CYTOSKELETON`)

Groups RHO-family GTPase, adhesion and actin-regulatory pathways that organise cell shape and movement.

Verified membership: 6 categories, 33 distinct core gene sets and 37 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| RAC1 signaling (`RAC1_SIGNALING`) | 10 | `PID_RAC1_PATHWAY`<br>`BIOCARTA_RAC1_PATHWAY` |
| RHOA-ROCK signaling (`RHOA_ROCK_SIGNALING`) | 11 | `PID_RHOA_PATHWAY`<br>`REACTOME_RHOA_GTPASE_CYCLE` |
| CDC42 signaling (`CDC42_SIGNALING`) | 8 | `PID_CDC42_PATHWAY`<br>`PID_CDC42_REG_PATHWAY` |
| PAK signaling (`PAK_SIGNALING`) | 2 | `KEGG_MEDICUS_REFERENCE_RAC_CDC42_PAK_ERK_SIGNALING_PATHWAY` |
| FAK/SRC adhesion signaling (`FAK_SRC_SIGNALING`) | 5 | `PID_FAK_PATHWAY` |
| Actin regulatory signaling (`ACTIN_REGULATORY_SIGNALING`) | 1 | `REACTOME_RHO_GTPASES_ACTIVATE_WASPS_AND_WAVES` |

### TGF-beta/SMAD signaling (`TGF_BETA_SMAD`)

Groups TGF-beta-family ligands and their SMAD-dependent developmental and homeostatic pathways.

Verified membership: 3 categories, 31 distinct core gene sets and 31 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| TGF-beta/SMAD signaling (`TGF_BETA_SMAD_SIGNALING`) | 17 | `KEGG_TGF_BETA_SIGNALING_PATHWAY`<br>`REACTOME_TGFBR3_REGULATES_TGF_BETA_SIGNALING` |
| BMP/SMAD signaling (`BMP_SMAD_SIGNALING`) | 8 | `PID_BMP_PATHWAY`<br>`REACTOME_SIGNALING_BY_BMP` |
| Activin/Nodal signaling (`ACTIVIN_NODAL_SIGNALING`) | 6 | `GOBP_NODAL_SIGNALING_PATHWAY`<br>`REACTOME_SIGNALING_BY_NODAL` |

### Wnt, Notch and Hedgehog signaling (`WNT_NOTCH_HEDGEHOG`)

Groups WNT, NOTCH and Hedgehog developmental signalling, including canonical and non-canonical branches.

Verified membership: 4 categories, 94 distinct core gene sets and 94 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| WNT/beta-catenin signaling (`WNT_BETA_CATENIN_SIGNALING`) | 37 | `WNT_SIGNALING`<br>`PID_BETA_CATENIN_DEG_PATHWAY`<br>`PID_BETA_CATENIN_NUC_PATHWAY` |
| Non-canonical WNT/PCP signaling (`NONCANONICAL_WNT_PCP_SIGNALING`) | 6 | `GOBP_NON_CANONICAL_WNT_SIGNALING_PATHWAY`<br>`REACTOME_PCP_CE_PATHWAY` |
| NOTCH signaling (`NOTCH_SIGNALING`) | 34 | `PID_NOTCH_PATHWAY`<br>`BIOCARTA_NOTCH_PATHWAY`<br>`GOBP_NOTCH_SIGNALING_PATHWAY` |
| Hedgehog signaling (`HEDGEHOG_SIGNALING`) | 17 | `KEGG_HEDGEHOG_SIGNALING_PATHWAY`<br>`PID_HEDGEHOG_2PATHWAY` |

### Hippo and mechanosignaling (`HIPPO_MECHANOSIGNALING`)

Groups Hippo/YAP/TAZ and mechanotransduction pathways that connect tissue mechanics to cell-state regulation.

Verified membership: 2 categories, 7 distinct core gene sets and 8 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Hippo/YAP/TAZ signaling (`HIPPO_YAP_TAZ_SIGNALING`) | 7 | `GOBP_HIPPO_SIGNALING`<br>`WP_HIPPO_SIGNALING_REGULATION` |
| Mechanotransduction signaling (`MECHANOTRANSDUCTION_SIGNALING`) | 1 | `WP_MECHANOREGULATION_AND_PATHOLOGY_OF_YAPTAZ_VIA_HIPPO_AND_NONHIPPO_MECHANISMS` |

### Cytokine/JAK/STAT signaling (`CYTOKINE_JAK_STAT`)

Groups cytokine pathways that transmit signals through JAK kinases and STAT transcription factors.

Verified membership: 6 categories, 52 distinct core gene sets and 52 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Type I interferon signaling (`TYPE_I_INTERFERON_SIGNALING`) | 10 | `WP_INTERFERON_TYPE_I_SIGNALING`<br>`GOBP_TYPE_I_INTERFERON_MEDIATED_SIGNALING_PATHWAY` |
| Type II interferon signaling (`TYPE_II_INTERFERON_SIGNALING`) | 6 | `WP_TYPE_II_INTERFERON_SIGNALING`<br>`GOBP_TYPE_II_INTERFERON_MEDIATED_SIGNALING_PATHWAY` |
| IL6/JAK/STAT3 signaling (`IL6_JAK_STAT3_SIGNALING`) | 9 | `WP_IL6_SIGNALING`<br>`BIOCARTA_IL6_PATHWAY` |
| IL2/STAT5 signaling (`IL2_STAT5_SIGNALING`) | 11 | `PID_IL2_STAT5_PATHWAY`<br>`WP_IL2_SIGNALING` |
| IL10/STAT3 signaling (`IL10_STAT3_SIGNALING`) | 5 | `BIOCARTA_IL10_PATHWAY` |
| IL12/IL23 signaling (`IL12_IL23_STAT4_3_SIGNALING`) | 11 | `PID_IL23_PATHWAY`<br>`BIOCARTA_IL12_PATHWAY` |

### Inflammatory NF-kB signaling (`INFLAMMATORY_NFKB`)

Groups canonical and non-canonical NF-kB pathways downstream of inflammatory receptors.

Verified membership: 4 categories, 54 distinct core gene sets and 55 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| TNF/NF-kB signaling (`TNF_NFKB_SIGNALING`) | 17 | `PID_TNF_PATHWAY`<br>`REACTOME_TNF_SIGNALING` |
| IL1/TLR-NF-kB signaling (`IL1_TLR_NFKB_SIGNALING`) | 17 | `PID_IL1_PATHWAY`<br>`WP_IL1_SIGNALING` |
| Canonical NF-kB signaling (`CANONICAL_NFKB_SIGNALING`) | 15 | `REACTOME_TRAF6_MEDIATED_NF_KB_ACTIVATION`<br>`GOBP_CANONICAL_NF_KAPPAB_SIGNAL_TRANSDUCTION` |
| Non-canonical NF-kB signaling (`NONCANONICAL_NFKB_SIGNALING`) | 6 | `REACTOME_TNFR2_NON_CANONICAL_NF_KB_PATHWAY`<br>`GOBP_NON_CANONICAL_NF_KAPPAB_SIGNAL_TRANSDUCTION` |

### Innate immune signaling (`INNATE_IMMUNE_SIGNALING`)

Groups pattern-recognition, nucleic-acid sensing and inflammasome pathways of innate immunity.

Verified membership: 4 categories, 54 distinct core gene sets and 54 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| TLR signaling (`TLR_SIGNALING`) | 30 | `BIOCARTA_TOLL_PATHWAY`<br>`GOBP_TOLL_SIGNALING_PATHWAY`<br>`PID_TOLL_ENDOGENOUS_PATHWAY` |
| RIG-I/MDA5 signaling (`RIG_I_MDA5_SIGNALING`) | 9 | `GOBP_RIG_I_SIGNALING_PATHWAY`<br>`KEGG_RIG_I_LIKE_RECEPTOR_SIGNALING_PATHWAY` |
| cGAS-STING signaling (`CGAS_STING_SIGNALING`) | 7 | `GOBP_CGAS_STING_SIGNALING_PATHWAY`<br>`KEGG_MEDICUS_REFERENCE_CGAS_STING_SIGNALING_PATHWAY` |
| NLRP3 inflammasome signaling (`NLRP3_INFLAMMASOME_SIGNALING`) | 8 | `REACTOME_THE_NLRP3_INFLAMMASOME`<br>`REACTOME_INFLAMMASOMES` |

### GPCR and second-messenger signaling (`GPCR_SECOND_MESSENGER`)

Groups GPCR and phospholipase pathways that use cyclic nucleotides, calcium and other second messengers.

Verified membership: 5 categories, 64 distinct core gene sets and 69 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| GPCR-cAMP-PKA signaling (`GPCR_CAMP_PKA_SIGNALING`) | 31 | `BIOCARTA_GPCR_PATHWAY`<br>`GOBP_CAMP_PKA_SIGNAL_TRANSDUCTION`<br>`KEGG_MEDICUS_REFERENCE_PKA_HOLOENZYME` |
| GPCR-PLC-calcium signaling (`GPCR_PLC_CALCIUM_SIGNALING`) | 16 | `BIOCARTA_GPCR_PATHWAY`<br>`KEGG_MEDICUS_REFERENCE_GPCR_PLCB_ITPR_SIGNALING_PATHWAY` |
| PLC-gamma signaling (`PLC_GAMMA_SIGNALING`) | 2 | `KEGG_MEDICUS_REFERENCE_TCR_PLCG_ITPR_SIGNALING_PATHWAY` |
| Calcium/calcineurin/NFAT signaling (`CALCIUM_CALMODULIN_CALCINEURIN`) | 12 | `REACTOME_CALCINEURIN_ACTIVATES_NFAT`<br>`BIOCARTA_NFAT_PATHWAY` |
| S1P/EDG receptor signaling (`S1P_EDG_SIGNALING`) | 8 | `WP_S1P_RECEPTOR_SIGNAL_TRANSDUCTION`<br>`PID_S1P_META_PATHWAY` |

### T- and B-cell receptor signaling (`T_CELL_B_CELL_RECEPTOR_SIGNALING`)

Groups antigen-receptor and costimulatory signalling in T and B lymphocytes.

Verified membership: 3 categories, 36 distinct core gene sets and 37 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| T-cell receptor signaling (`TCR_SIGNALING`) | 17 | `GOBP_T_CELL_RECEPTOR_SIGNALING_PATHWAY`<br>`KEGG_T_CELL_RECEPTOR_SIGNALING_PATHWAY` |
| B-cell receptor signaling (`BCR_SIGNALING`) | 13 | `WP_B_CELL_RECEPTOR_SIGNALING`<br>`GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY` |
| CD28 costimulation signaling (`CD28_COSTIMULATION_SIGNALING`) | 7 | `REACTOME_CD28_DEPENDENT_VAV1_PATHWAY`<br>`REACTOME_CO_STIMULATION_BY_CD28` |

### Death-receptor signaling (`DEATH_RECEPTOR_SIGNALING`)

Groups extrinsic cell-death pathways initiated by death receptors and their proximal signalling complexes.

Verified membership: 2 categories, 14 distinct core gene sets and 16 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| FAS/death receptor signaling (`FAS_DEATH_RECEPTOR_SIGNALING`) | 10 | `REACTOME_DEATH_RECEPTOR_SIGNALING`<br>`PID_FAS_PATHWAY` |
| TRAIL death receptor signaling (`TRAIL_DEATH_RECEPTOR_SIGNALING`) | 6 | `REACTOME_DEATH_RECEPTOR_SIGNALING`<br>`PID_TRAIL_PATHWAY` |

### Nuclear-receptor signaling (`NUCLEAR_RECEPTOR_SIGNALING`)

Groups signalling and transcriptional programmes controlled by steroid and metabolic nuclear receptors.

Verified membership: 3 categories, 22 distinct core gene sets and 22 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Estrogen receptor signaling (`ESTROGEN_RECEPTOR_SIGNALING`) | 9 | `WP_ESTROGEN_RECEPTOR_PATHWAY`<br>`GOBP_ESTROGEN_RECEPTOR_SIGNALING_PATHWAY` |
| Androgen receptor signaling (`ANDROGEN_RECEPTOR_SIGNALING`) | 5 | `WP_ANDROGEN_RECEPTOR_SIGNALING` |
| PPAR signaling (`PPAR_SIGNALING`) | 8 | `WP_PPAR_SIGNALING`<br>`KEGG_PPAR_SIGNALING_PATHWAY` |

### Stress-response signaling (`STRESS_RESPONSE_SIGNALING`)

Groups NRF2 antioxidant and unfolded-protein-response branches that manage oxidative and endoplasmic-reticulum stress.

Verified membership: 4 categories, 24 distinct core gene sets and 26 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| NRF2/KEAP1 signaling (`NRF2_KEAP1_SIGNALING`) | 11 | `WP_NRF2_PATHWAY`<br>`KEGG_MEDICUS_REFERENCE_KEAP1_NRF2_SIGNALING_PATHWAY` |
| PERK/ATF4 UPR signaling (`UPR_PERK_ATF4_SIGNALING`) | 8 | `KEGG_MEDICUS_VARIANT_SCRAPIE_CONFORMATION_PRPSC_TO_PERK_ATF4_SIGNALING_PATHWAY`<br>`KEGG_MEDICUS_VARIANT_MUTATION_CAUSED_ABERRANT_SNCA_TO_PERK_ATF4_SIGNALING_PATHWAY` |
| IRE1/XBP1 UPR signaling (`IRE1_XBP1_SIGNALING`) | 3 | `GOBP_IRE1_MEDIATED_UNFOLDED_PROTEIN_RESPONSE` |
| ATF6 UPR signaling (`ATF6_UPR_SIGNALING`) | 4 | `GOBP_ATF6_MEDIATED_UNFOLDED_PROTEIN_RESPONSE` |

### RTK, matrix and guidance signaling (`RTK_MATRIX_GUIDANCE_SIGNALING`)

Groups receptor and guidance pathways that connect extracellular cues, matrix interactions and directed cell behaviour.

Verified membership: 3 categories, 20 distinct core gene sets and 20 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| Ephrin/EPH signaling (`EPHRIN_EPH_SIGNALING`) | 11 | `REACTOME_EPH_EPHRIN_SIGNALING`<br>`REACTOME_EPHRIN_SIGNALING` |
| DDR1/DDR2 collagen receptor signaling (`DDR1_DDR2_COLLAGEN_RECEPTOR_SIGNALING`) | 1 | `GOBP_COLLAGEN_ACTIVATED_TYROSINE_KINASE_RECEPTOR_SIGNALING_PATHWAY` |
| Semaphorin/Plexin/Neuropilin signaling (`SEMAPHORIN_PLEXIN_NEUROPILIN_SIGNALING`) | 8 | `GOBP_SEMAPHORIN_PLEXIN_SIGNALING_PATHWAY`<br>`REACTOME_SEMAPHORIN_INTERACTIONS` |

### Chemokine GPCR signaling (`CHEMOKINE_GPCR_SIGNALING`)

Groups chemokine-receptor and related GPCR signalling axes that guide cell positioning and communication.

Verified membership: 1 category, 12 distinct core gene sets and 12 category assignments.

| Category | Core members | Representative member gene sets |
|---|---:|---|
| CXCR4/CXCL12 signaling (`CXCR4_CXCL12_SIGNALING`) | 12 | `GOBP_CXCL12_ACTIVATED_CXCR4_SIGNALING_PATHWAY`<br>`PID_CXCR4_PATHWAY` |

# Use a category in study evidence

The catalogue describes the dictionary. Study-specific NES, adjusted
P values and leading edges are available in the report's category evidence.
In a single-analysis report, choose the analysis and collection, then select
**Category evidence** in **Section** and the exact category in **Category**.
See the [report guide](scientific-report.html) for the complete navigation and
download distinctions.
For custom resources and schema requirements, see
[Custom dictionaries and supercategories](custom-dictionaries-and-supercategories.html).
