---
title: "Explore a saved report with Shiny"
lang: en-GB
output:
  rmarkdown::html_vignette:
    mathjax: null
    includes:
      in_header: vignette-layout.inc
vignette: >
  %\VignetteIndexEntry{Explore a saved report with Shiny}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use Shiny to browse a completed lisaR report and generate selected figures from
its saved results. The application does not repeat differential expression,
GSEA or ORA. It writes generated material to a separate exploration workspace
and leaves the original scientific report unchanged.

To inspect existing pages and downloads without generating anything, open the
saved report's `report_index.html` directly. That route needs neither R nor
Shiny.

## Open the saved analysis

Install the optional application packages once in the R library where lisaR is
installed:

```{r install-shiny, eval=FALSE}
install.packages(c("shiny", "callr", "ps"))
```

`shiny` provides the interface. `callr` runs a figure worker separately so
you can continue to browse. `ps` supports worker status checks. Some figures
need further packages or saved inputs; the application reports when one is not
available for the selected run.

Set `run_dir` to the completed report folder and open it:

```{r open-saved-run, eval=FALSE}
library(lisaR)

run_dir <- "path/to/your/saved-run"
explore_lisa_run(run_dir)
```

After [Quick Start](quick-start.html), you can instead use
`run_dir <- result$output_dir` in the same R session.

In an environment where R must not launch the browser, use the supported
option below and open the loopback URL printed by Shiny in the browser
available to that environment:

```{r open-without-browser-launch, eval=FALSE}
explore_lisa_run(run_dir, launch.browser = FALSE)
```

This option leaves the application running while you open its address yourself.

## Choose one exact item

The saved report opens inside a Shiny shell. To reach its figure controls:

1. Open the analysis or contrast, then choose the collection.
2. Under **Section**, select **Category evidence** and follow **Explore a category**.
3. Choose the category in the evidence sheet's selector.

For Quick Start, use **Synthetic response A**, **GOBP-C2** and
**Synthetic immune signalling**. After the custom-dictionary tutorial, use the
same analysis and collection with **Example of signal response**.

Use **Explore a category** for this route. Clicking a category label in an
overview graph opens the static evidence page outside the Shiny shell.

Each button names one product. Examples include **Generate volcano**,
**Generate gene card**, **Generate gene heatmap** and
**Generate KEGG pathway map**. Contrast pages can show named buttons such as
**Generate paired gene heatmap** or **Generate gene-category network**.

Where a product has another selector, choose one exact value before pressing
its button. For a gene figure, check the gene. For a heatmap, check the offered
scale. For a native KEGG map, check both the pathway name and identifier.
Changing a selection, opening a category or searching for a gene does not
generate a figure.

A native KEGG map also requires an authorised immutable local snapshot and
the corresponding saved run information. KEGG-derived TERM2GENE membership
can support enrichment but does not supply a native pathway diagram. See
[Resources and provenance](resources-and-provenance.html).

## Generate the figure and follow its state

Press the named **Generate** button once. The request moves through the states
that apply to it:

| State | What it means |
|---|---|
| `ungenerated` | No request has been submitted for this figure. |
| `queued` | The request is saved and waiting for the single figure-worker slot. |
| `running` | A separate R worker is producing this figure from saved results. |
| `available` | The generated files passed their checks and can be reused. |
| `not_applicable` | The saved run lacks an input or condition required by this product. |
| `failed` | The request ended without an available figure; read its message. |

A second background figure request can remain `queued` behind the one that is
`running`. You do not need to close the report while it waits. Generating a
figure is presentation work over saved results, not a new enrichment analysis.

## Inspect, download and reuse the result

When the state is `available`, inspect the image in the same category sheet.
Use whichever PNG, PDF, source-data or R-script download links that product
actually provides. Keep the source data and settings with a downloaded recipe.

Reopening the same saved analysis reuses a generated figure while its source
run, settings and files still validate. Opening the report never generates a
missing figure automatically.

## Resolve unavailable or failed requests

`not_applicable` is a product-availability result, not a negative biological
finding. Read the explanation. For example, a gene heatmap may need a saved
expression matrix, while a native KEGG map needs the authorised snapshot
described above. Other saved report sections remain available.

For a `failed` request, keep its message and correct the named input, package
or resource outside the immutable source run. The same named **Generate**
button becomes available again; there is no separate **Retry** button.

## Cancel a figure through the advanced R API

The Shiny shell has no visible per-figure cancel button. If you submit figures
through R, keep the workspace and the `job_id` returned by
`lisa_explore_submit()`:

```{r cancel-figure-api, eval=FALSE}
ws <- lisa_explore_open(run_dir)
request <- lisa_figure_request(
  unit_type = "single_de",
  analysis_id = "response_a",
  collection = "GOBP-C2",
  category_id = "SYN_SIGNAL",
  product = "volcano"
)
job <- lisa_explore_submit(ws, request, background = TRUE)

lisa_explore_status(ws)
lisa_explore_cancel(ws, job$job_id)
```

The identifiers above belong to the synthetic sample project. Use IDs from
your saved report for another study.

A background request can queue behind another figure. A foreground API request
can instead be refused with an instruction to wait or submit it in the
background. Follow that message rather than restarting R.

A queued job is cancelled immediately and recorded as `failed` with the
cancellation reason. A running worker records the request, continues until its
next safe point and only then reaches a terminal `failed` state. Do not close
or stop the process to force cancellation. This API route is separate from the
export control below.

## Save an expanded report

1. In **Save expanded report to**, enter a new folder outside the original
   report. The destination must not already exist.
2. Press **Save expanded report**.
3. Wait while the toolbar shows **Export queued** or **Export running**. Continue
   only when it reports **Saved expanded report to** the chosen destination.
4. Copy or zip the whole new folder. At the destination, open its
   `report_index.html` and keep every subfolder with it.

The export takes a snapshot of figures that are `available` when saving
begins. It does not generate queued, running or ungenerated figures. If a
figure worker prevents the export from starting, wait for the displayed figure
state to finish, then press **Save expanded report** again.

If the toolbar reports **Export failed**, keep the message and do not move a
partial staging folder. Correct the named problem and choose a new destination
before trying again.

Use **Cancel export** only for an export that is queued or running. A queued
export cancels immediately. A running export stops at its next safe point, so
wait for the displayed terminal message before closing R.

The expanded folder contains the original report and the available figures
captured by that export. It opens without R or Shiny, with working figures and
downloads but no live generation controls. See
[Read and share a saved report](scientific-report.html) for navigation.
